### 2023-07-18 paper/appendix/code.tex
Added LaTeX paper source code

\section{Appendix -- Application \& Code Details}
\subsection{Application}
Hello World - Application \& Code Details

\subsection{Code (Python)}

\clearpage

### 2023-07-18 paper/appendix/dataset.tex
Added LaTeX paper source code

\section{Appendix -- Dataset Details}
The following details in this appendix are for clarification purposes.
All examples in this appendix are created via PySpark using dataframes.

\subsection{Using PySpark to Load Data}
All example code was created using Jupyter Notebooks creating a PySpark session. The deails of setting up such session are below in the code example. All subsequent details are derived from the \texttt{sc}\footnote{sc = Spark Session}.
\begin{minted}{python}
# Setup Spark Session
# Spark Session Name: Qualyfing_Exam
from pyspark.sql import SparkSession

# Get Imports Needed
from pyspark.sql.functions import col, udf

# Get Datatypes needed for DataFrame manipulation
from pyspark.sql.types import IntegerType, StringType

# Setup Spark Session
sc = SparkSession \
        .builder \
        .master("local[*]") \
        .appName("Qualyfing_Exam") \
        .getOrCreate()
\end{minted}

\subsection{Details on English Dataset}
The English dataset based on tweets\footnote{Tweets from Twitter} are saved as a \texttt{csv}\footnote{csv = Comma Separated Values} file. The following are some simply details on the laod and processing of such file.
\begin{minted}{python}
# Read csv file to Saprk dataframe
df_english = sc.read.csv("data/english.csv", header=True, inferSchema=True)
# Group by a column and count the occurrences in df_english
count_english = df_english.groupBy('sentiment').count()

# Show the result for df_english
count_english.show()
+----------+-----+
| sentiment|count|
+----------+-----+
|   boredom|  179|
|      love| 3842|
|    relief| 1526|
|       fun| 1776|
|      hate| 1323|
|   neutral| 8638|
|     anger|  110|
| happiness| 5209|
|  surprise| 2187|
|   sadness| 5165|
|     worry| 8459|
|enthusiasm|  759|
|     empty|  827|
+----------+-----+

\end{minted}

\subsection{Details on German Dataset}
The German dataset was downloaded from the ETH's\footnote{ETH (German: Eidgenössische Technische Hochschule) or Swiss Federal Institute of Technology} servers via "link" as a \texttt{JSON} file.
\begin{minted}{python}
# Read JSON file to Saprk dataframe
df_german = sc.read.json("data/german.json")

# Group by a column and count the occurrences in df_german
count_german = df_german.groupBy('article_emotion').count()

# Show the result for df_german
count_german.show(truncate=0)
+----------------------------------+-----+
|article_emotion                   |count|
+----------------------------------+-----+
|[Ärger, Antizipation, Traurigkeit]|1    |
|[Überraschung, Vertrauen]         |14   |
|[Überraschung, Freude]            |15   |
|[Ärger, Vertrauen]                |7    |
|[Ekel, Traurigkeit]               |4    |
|[Keine, Antizipation]             |5    |
|[Vertrauen]                       |130  |
|[Freude]                          |58   |
|[Ärger, Antizipation]             |31   |
|[Ärger]                           |114  |
|[Überraschung]                    |154  |
|[Antizipation, Traurigkeit]       |15   |
|[Antizipation, Freude, Vertrauen] |2    |
|[Vertrauen, Angst]                |5    |
|[Traurigkeit]                     |101  |
|[Antizipation]                    |413  |
|[Traurigkeit, Vertrauen]          |7    |
|[Antizipation, Vertrauen]         |104  |
|[Ärger, Traurigkeit, Angst]       |1    |
|[Unklar]                          |314  |
+----------------------------------+-----+

\end{minted}

\clearpage

### 2023-07-18 paper/main.tex
Added LaTeX paper source code

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{fullpage}
\usepackage{multirow}

\usepackage{minted}
\setminted[]{linenos, samepage, numbersep=5pt, frame=lines}

\usepackage{parskip}
\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\usepackage[affil-it]{authblk}
\usepackage{amsfonts, amssymb, amsmath}

\usepackage{wrapfig}
\usepackage{graphicx}
\graphicspath{ {./images/} }

\usepackage{float}

\usepackage[backend=biber]{biblatex}
\addbibresource{sources.bib}
% Use the followign Command to validate *.bib file: biber --tool --validate-datamodel qualifying_exam_paper.bib

\title{Improving Emotion Detection through Real-Time Translation of Text to ML Models Trained in Different Languages}

\author{Richard Hoehn%
	\thanks{Electronic address: \texttt{rhoehn@mtmail.mtsu.edu}; corresponding author}}
\affil{Middle Tennessee State University}

\author{Dr. Jaishree Ranganathan%
	\thanks{Electronic address: \texttt{jaishree.ranganathan@mtsu.edu}}}
\affil{Middle Tennessee State University}

\begin{document}

\maketitle

\begin{abstract}
This Qualifying Exam\footnote{PhD students of MTSU Computational and Data  Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages. The research tries to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis.

By extending an original language dataset with a translated dataset from another language and utilizing parallel processing with PySpark, this research project aims to achieve higher prediction rates in ED. The results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper demonstrate the effectiveness of such an approach and how the real-time translation technique can enhance emotion detection capabilities.
\end{abstract}
\clearpage

\tableofcontents
\clearpage

\section{Introduction}
\subsection{Background on Human Emotion Theoretical Framework}

\begin{wrapfigure}{r}{0.5\textwidth} %this figure will be at the right
    \centering
    \includegraphics[width=0.5\textwidth]{Plutchik-Wheel}
    \caption{Plutchik Wheel of Emotions}
    \label{fig:plutchik-wheel}
\end{wrapfigure}

The concept of the eight main emotions can vary depending on the theoretical framework or model being considered. One well-known model is the Plutchik's Wheel of Emotions figure \ref{fig:plutchik-wheel}, which proposes eight primary emotions and their contrasting pairs. Here are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp}:

\begin{itemize}
\item Joy: A feeling of happiness, contentment, or delight.
\item Sadness: A state of unhappiness, sorrow, or grief.
\item Anger: An intense emotional response associated with feelings of displeasure, frustration, or hostility.
\item Fear: An emotional response triggered by perceived threats or danger, often associated with anxiety or terror.
\item Surprise: A sudden and unexpected reaction to something unusual or startling.
\item Disgust: A strong aversion or revulsion towards something unpleasant, offensive, or repulsive.
\item Trust: A sense of reliance, confidence, or faith in someone or something.
\item Anticipation: A feeling of excited expectation or eagerness towards future events.
\end{itemize}
It's important to note that emotions are complex and multifaceted, and different models may propose alternative categorizations or variations in the number of primary emotions. However for the scope of this research project the most common 7-8 emotions are used as described above.

\subsection{Background and significance of Emotion Detection (ED) analysis in Machine Learning models}
Emotion Detection (ED) analysis in Machine Learning models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific target\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions to name a few.

By accurately identifying and understanding emotions from text data, ML models can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.

\begin{quote}
    I usually know almost exactly how I feel. The problem is, I just can’t tell anyone.
    \flushright -- Meg Cabot
\end{quote}

Due to the before mentioned this field of research provides significance in enabling machines to comprehend and respond appropriately to human emotions, contributing to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications that are growing rapidly.

\subsection{ED challenges, including lack of comprehensive labeled datasets and language fragmentation}
ED is growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes is much smaller. The primary reason of the lack of data stems from the emotion detection requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created by humans. To compound on the issue of the lack of data; the many datasets that are available are in multiple languages.

\subsection{Motivation for the research project focusing on extending ED datasets through translation and evaluating the impact on prediction rates}
With the before mentioned issues regarding the quantity/lack and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED. The hope is that by use of translation that the predictability of ED can be improved by use of real- real-time language conversion and processing.

\subsection{Objectives \& Scope of the Research}
The project is comprised of three main points / goals which are listed in following sub-sections with details in each point as follows:

\subsubsection{Data Collection / Translation / ML Model Training}

\begin{wrapfigure}{r}{0.4\textwidth} %this figure will be at the right
    \centering
    \includegraphics[width=0.4\textwidth]{Dataset-Extension}
    \caption{Example of English dataset extension by translating from German.}
    \label{fig:dataset-extension}
\end{wrapfigure}

The procurement of two ED datasets is needed. The first will be in English and the second in German. Each dataset will be translated in the other’s language for extending purposes. As an example figure \ref{fig:dataset-extension}, the English will be translated to German and the German to English by which the original datasets are extended.

Once the two datasets have been translated, we now extend the original language dataset with the translated once. In essence extending the datasets to create a larger one. With both extended datasets we can now train our ML models for English and German.

\subsubsection{PySpark Application Development for Translation \& Prediction}
A program will be built in Jupiter notebooks that uses translations APIs to convert an English sentence to German and the same if a German sentence is fed to English. Both sentences are then processed in real-time on both the English and German models for the Emotion Detection and compared. The average of the prediction will be used for final prediction of the emotion that was processed.

\subsubsection{Written \& Oral Presentation of Research Finding}
In conclusion of the project a research paper and oral presentation is required to pass MTSU Computation Science PhD Qualifying Exam. Such paper is being presented with this paper and the oral presentation will be in Microsoft Power Point format.

The listing the datasets used, code for the application, and details of the research project process in this research paper is openly available and hosted on a GitHub repository, allowing for easy access, reproducible, and collaborative engagement with the research findings. The repository is open and available under: \texttt{https://github.com/richardhoehn/mtsu.coms.qualifying-exam}\cite{Hoehn_Improving_Emotion_Detection_2023}

\section{Literature Review}

\subsection{Emotion Detection Importance, Data Limitations, Diverse Populations}
Emotion Detection (ED) holds significant importance in various domains, including psychology, social sciences, and human-computer interaction. It enables the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states. However, the field faces limitations due to the scarcity of comprehensive labeled datasets. The lack of diverse and large-scale data hinders the development and training of accurate ED models, limiting their effectiveness and general ability across different contexts and populations.

In recent research, efforts have been made to address these limitations by exploring new techniques for emotion detection. For instance, a study by Shalini Kapoor, Tarun Kumar \cite{KAPOOR2023120882} proposes analyzing the entire spectrum of the language source to predict emotional changes.

Overcoming the scarcity of labeled datasets and embracing innovative approaches like the before mentioned study by Shalini Kapoor, Tarun Kumar \cite{KAPOOR2023120882} and their analysis can contribute to the advancement of ED research. Such advancements hold the potential to enhance the accuracy and applicability of emotion detection models, enabling a deeper understanding of human emotions and more effective responses in various domains.

Emotions can differ across age groups, genders, cultures, and languages. Including data from diverse populations helps in developing inclusive and culturally sensitive ED models. It ensures that the models are not biased towards specific groups and can accurately detect emotions in a wide range of individuals.

\subsection{Challenges by the lack of Labeled Datasets and the Fragmentation caused by Different Languages}
Unlike images language operate very differently, therefore the lack of labeled datasets poses a significant challenge in text-based Emotion Detection (ED) analysis as it limits the availability of reliable training data for ML models. This scarcity hampers the model's ability to learn and generalize emotions effectively. Additionally, the fragmentation caused by different languages further exacerbates the issue, as it reduces the size and diversity of data available for training, resulting in limited cross-lingual generalization and potentially biased models. Overcoming these challenges is crucial for advancing ED research and improving the accuracy and applicability of emotion detection models across various languages and cultures.

\section{Methodology}

\subsection{Data procurement process, including obtaining ED datasets in English and German}
The data procurement was relatively straight forward and in essence entailed using Google search terms for the Emotion Detection in English and German. Some of the main German data was obtained from the dataset built by ETH's Emotion and Stance Detection for German Text \cite{mascarell-etal-2021-stance} which also hosts a good website with their finds \footnote{https://mtc.ethz.ch/research/natural-language-processing/emotion-stance.html}. The English dataset was downloaded from Kaggle \footnote{https://www.kaggle.com/datasets/pashupatigupta/emotion-detection-from-text} based on Tweets collected in 2021. The English dataset contains over 38,000 tweets and their corresponding emotion labeled. Similarly the German contains over 12,000 sentences and their respective emotions associated.

An important part of the process is to make sure we match the English emotions to those of the German language emotions. Since the tow languages are dissimilar we opted to match exact emotions to each other that can be seen in Table \ref{table:dataset_labels} below.

\begin{table}[h!]
\centering
\begin{tabular}{ | c c | c c | c | }
    \hline
    \multicolumn{5}{|c|}{Emotion Datasets Labels} \\

    \hline
    \multicolumn{2}{|c|}{English} & \multicolumn{2}{|c|}{German} & \multirow{2}{*}{Used} \\
    Name & Count & Name & Count \\
    \hline
    Boredom    &  179 & ---           & --- & NO  \\
    Love       & 3842 & Vertrauen     & 130 & YES \\
    Relief     & 1526 & ---           & --- & NO  \\ 
    Fun        & 1776 & ---           & --- & NO  \\
    Hate       & 1323 & Ekel          &   4 & YES \\
    Neutral    & 8638 & Unklar        & 314 & YES \\
    Anger      &  110 & Ärger         & 114 & YES \\
    Happiness  & 5209 & Freude        &  58 & YES \\
    Surprise   & 2187 & Überraschung  & 154 & YES \\
    Sadness    & 5165 & Traurigkeit   & 101 & YES \\
    Worry      & 8459 & Angst         &   6 & YES \\
    Enthusiasm &  759 & Antizipation  & 413 & YES \\
    Empty      &  827 & ---           & --- & NO  \\
    \hline
\end{tabular}
\caption{Dataset of English \& German and their respective label counts.}
\label{table:dataset_labels}
\end{table}

\subsection{Development of a translation application for converting different languages into English and vice versa}
Translation of the emotion datasets was performed with Google Translator based on the Medium article\cite{Nidhaloff_how_to_translate_text_with_python} in which the author describes option fro simple API translations. As an example the following is a brief introduction to Google Translator's API:

\begin{quote}
\begin{minted}{python}
# Using Deep Translator to leveage Google Translate
# Link: https://cloud.google.com/translate/docs/reference/libraries/v2/python
from deep_translator import GoogleTranslator

sentence = 'Chocolate milk is so much better through a straw.'

translated = GoogleTranslator(source='auto', target='de').translate(sentence)
print(translated) # Schokoladenmilch schmeckt durch einen Strohhalm viel besser.

translated_back = GoogleTranslator(source='auto', target='en').translate(translated)
print(translated_back) # Chocolate milk tastes much better through a straw.
\end{minted}
\end{quote}

\subsection{Building and training multiple ML models using the extended datasets and parallel processing approach with PySpark}
\ldots

\subsection{Evaluation Metrics and methodologies for measuring the performance of the ML models}
The evaluation of ML models trained on text datasets is a critical step in assessing their performance and reliability. Our research project is no different and the following assumptions are used to determine the effectiveness of the methodology of extending our datasets:
\begin{itemize}
\item \textbf{Accuracy:} Accuracy measures the overall correctness of the emotion predictions made by the model. It is calculated as the ratio of correctly classified instances to the total number of instances in the dataset. However, accuracy may not provide a complete picture, especially if the dataset is imbalanced.
\item \textbf{Precision, Recall, and F1 Score:} Precision represents the proportion of correctly predicted emotions among all instances predicted as a specific emotion. Recall measures the proportion of correctly predicted emotions among all instances that truly belong to a specific emotion. F1 score combines precision and recall into a single metric, providing a balanced evaluation.
\item \textbf{Cross-Validation:} Cross-validation techniques, such as k-fold or stratified cross-validation, divide the dataset into multiple subsets for training and testing. This helps evaluate the model's performance across different data samples, mitigating over-fitting and providing a more robust estimation of its generalization ability.
\item \textbf{Error Analysis:} Performing error analysis on the model's predictions helps identify specific patterns or instances where the model might struggle or make frequent mistakes. This analysis can guide improvements and provide insights for future model development.
\end{itemize}

When evaluating such models, several factors come into play. First and foremost, the choice of appropriate evaluation metrics is crucial, depending on the specific task at hand, such as text classification, sentiment analysis, or named entity recognition. Metrics like accuracy, precision, recall, F1 score, or area under the receiver operating characteristic curve (AUC-ROC) are commonly used to measure the models' performance.

\section{Results and Analysis}

\subsection{Presentation of the prediction results obtained from the extended datasets}

\ldots

\subsection{Comparison of Prediction Result of the Original and Extended Language (translated) Datasets}
\ldots

\subsection{Analysis of the impact of Extending Datasets by Translation and Performance of the Combination of ML models}
\ldots

\subsection{Discussion of the findings and their implications for improving Emotion Detection}
\ldots

\section{Demonstration and Application}

\subsection{Description of the developed translation application for real-time text translation}
Details of the entire application can be found on the github repository \cite{Hoehn_Improving_Emotion_Detection_2023}.

\subsection{Showcase of the application's functionality and its integration with the Emotion Detection process}
\ldots

\subsection{Demonstration of the real-time processing of translated text and emotion detection using the ML models}
\ldots

\subsection{GIT Repository Details}
The listing the datasets used, code for the application, and details of the research project process in this research paper is openly available and hosted on a GitHub repository, allowing for easy access, reproducible, and collaborative engagement with the research findings. The repository is open and available under: \texttt{https://github.com/richardhoehn/mtsu.coms.qualifying-exam}\cite{Hoehn_Improving_Emotion_Detection_2023}
\clearpage

\addcontentsline{toc}{section}{References}
\printbibliography
\clearpage

\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\clearpage

% Appendix Details
\appendix
\input{appendix/dataset} % Note: SubSection headers are handled within the *.tex file 
\input{appendix/code} % Note: SubSection headers are handled within the *.tex file

\end{document}

### 2023-07-18 paper/appendix/code.tex
Paper Updates

@@
 \section{Appendix -- Application \& Code Details}
+Hello World - Application \& Code Details
 \subsection{Application}
-Hello World - Application \& Code Details
+\label{appendix:code_application}
+Hello World - Application
+\clearpage
 
-\subsection{Code (Python)}
+\subsection{Code Details}
+Hello World - Code Details
+
+\subsection{RESTful API Server}
+\label{appendix:code-api-server}
+The RESTful\footnote{RESTful API is an interface that two computer systems use to exchange information securely over the internet.} API Server will be a simple Python Flask server listening on port \texttt{8080} on the local machine.
 

### 2023-07-18 paper/appendix/dataset.tex
Paper Updates

@@
 All example code was created using Jupyter Notebooks creating a PySpark session.
-The deails of setting up such session are below in the code example.
+The details of setting up such session are below in the code example.
 All subsequent details are derived from the \texttt{sc}\footnote{sc = Spark Session}.
-\begin{minted}{python}
+\begin{minted}[fontsize=\small]{python}
+# General Imports
+import glob
+import shutil
+
 # Setup Spark Session
-# Spark Session Name: Qualyfing_Exam
 from pyspark.sql import SparkSession
 
-# Get Imports Needed
-from pyspark.sql.functions import col, udf
+# Get Spark Functions Needed
+from pyspark.sql.functions import col, udf, split, explode
 
@@
         .master("local[*]") \
-        .appName("Qualyfing_Exam") \
+        .appName("data_clean_up") \
         .getOrCreate()
+
+# Print Spark Version being run
+print("Spark V: ", sc.version)
 \end{minted}
+\clearpage
 
 \subsection{Details on English Dataset}
-The English dataset based on tweets\footnote{Tweets from Twitter} are saved as a \texttt{csv}\footnote{csv = Comma Separated Values} file.
-The following are some simply details on the laod and processing of such file.
-\begin{minted}{python}
-# Read csv file to Saprk dataframe
+\label{appendix:dataset_english}
+The English dataset is based on tweets\footnote{Tweets from Twitter} that are saved as a \texttt{csv}\footnote{csv = Comma Separated Values} file.
+The following code was use to filter and process the file for use in this project.
+\begin{minted}[fontsize=\small]{python}
+# *********************************
+# *** English Data Preparations ***
+# *********************************
+
+# Load English CSV File into a Dataframe
 df_english = sc.read.csv("data/english.csv", header=True, inferSchema=True)
-# Group by a column and count the occurrences in df_english
-count_english = df_english.groupBy('sentiment').count()
+print(f'Count (raw): {df_english.count()}')
 
-# Show the result for df_english
-count_english.show()
-+----------+-----+
-| sentiment|count|
-+----------+-----+
-|   boredom|  179|
-|      love| 3842|
-|    relief| 1526|
-|       fun| 1776|
-|      hate| 1323|
-|   neutral| 8638|
-|     anger|  110|
-| happiness| 5209|
-|  surprise| 2187|
-|   sadness| 5165|
-|     worry| 8459|
-|enthusiasm|  759|
-|     empty|  827|
-+----------+-----+
+# Print Columns
+print(f'English Source Columns: {df_english.columns}\n')
 
+# Filter only lavbels that we mathc with the German counter parts
+filter_values = ["love", "hate", "neutral", "anger", "happiness", "surprise", "sadness", "worry", "enthusiasm"]
+df_english_filtered = df_english.filter(col("sentiment").isin(filter_values))
+
+# Remove unnecessary column
+df_english_filtered = df_english_filtered.drop("tweet_id")
+
+# Rename Columns
+df_english_filtered = df_english_filtered.withColumnRenamed("sentiment", "emotion")
+df_english_filtered = df_english_filtered.withColumnRenamed("content", "sentence")
+
+# Make sure the column order to the same for both german and english csv files
+df_english_filtered = df_english_filtered.select('sentence', 'emotion')
+
+print(f'Count (filtered): {df_english_filtered.count()}')
+
+# Group By for Details & Count
+df_english_grouped = df_english_filtered.groupBy('emotion').count()
+
+# Show Groupings and Respetive Counts
+print('\nGrouped & Count by "emotion"')
+df_english_grouped.show(truncate=0)
+
+# Save Dataframe to CSV
+directory_path = 'data/spark_data_parts'
+df_english_filtered.coalesce(1).write.csv(directory_path, header=True, mode="overwrite")
+
+file_pattern = 'part-00000*.csv'
+file_path = glob.glob(directory_path + '/' + file_pattern)[0]
+
+shutil.move(file_path, './data/data_en.csv')
 \end{minted}
+\clearpage
 
 \subsection{Details on German Dataset}
+\label{appendix:dataset_german}
 The German dataset was downloaded from the ETH's\footnote{ETH (German: Eidgenössische Technische Hochschule) or Swiss Federal Institute of Technology} servers via "link" as a \texttt{JSON} file.
-\begin{minted}{python}
-# Read JSON file to Saprk dataframe
+Since the the file was \texttt{JSON} we will need to filter and explode the file based on teh emotions that are attached to it.
+\begin{minted}[fontsize=\small]{python}
+# ********************************
+# *** German Data Preparations ***
+# ********************************
+
+# Load German JSON File into a Dataframe
 df_german = sc.read.json("data/german.json")
+print(f'Count (raw): {df_german.count()}')
 
-# Group by a column and count the occurrences in df_german
-count_german = df_german.groupBy('article_emotion').count()
+# Print Columns
+print(f'German Source Columns: {df_german.columns}\n')
 
-# Show the result for df_german
-count_german.show(truncate=0)
-+----------------------------------+-----+
-|article_emotion                   |count|
-+----------------------------------+-----+
-|[Ärger, Antizipation, Traurigkeit]|1    |
-|[Überraschung, Vertrauen]         |14   |
-|[Überraschung, Freude]            |15   |
-|[Ärger, Vertrauen]                |7    |
-|[Ekel, Traurigkeit]               |4    |
-|[Keine, Antizipation]             |5    |
-|[Vertrauen]                       |130  |
-|[Freude]                          |58   |
-|[Ärger, Antizipation]             |31   |
-|[Ärger]                           |114  |
-|[Überraschung]                    |154  |
-|[Antizipation, Traurigkeit]       |15   |
-|[Antizipation, Freude, Vertrauen] |2    |
-|[Vertrauen, Angst]                |5    |
-|[Traurigkeit]                     |101  |
-|[Antizipation]                    |413  |
-|[Traurigkeit, Vertrauen]          |7    |
-|[Antizipation, Vertrauen]         |104  |
-|[Ärger, Traurigkeit, Angst]       |1    |
-|[Unklar]                          |314  |
-+----------------------------------+-----+
+# We Need to "split" the "artice_emotion" column since the ETH team listed
+# multiple emotions in one column
+# In order to "explode" the column to it's distinct rows
+df_german_exploded = df_german.select('*', explode('article_emotion').alias('emotion'))
 
+# Rename Column
+df_german_exploded = df_german_exploded.withColumnRenamed("title", "sentence")
+
+# Remove unnecessary column
+df_german_exploded = df_german_exploded.drop("article_id")
+df_german_exploded = df_german_exploded.drop("article_stance")
+df_german_exploded = df_german_exploded.drop("paragraphs")
+df_german_exploded = df_german_exploded.drop("source")
+df_german_exploded = df_german_exploded.drop("article_emotion")
+df_german_exploded = df_german_exploded.drop("snippet")
+
+# Make sure the column order to the same for both german and english csv files
+df_german_exploded = df_german_exploded.select('sentence', 'emotion')
+
+print(f'Count (filtered): {df_german_exploded.count()}')
+
+# Group By for Details & Count
+df_german_grouped = df_german_exploded.groupBy('emotion').count()
+
+# Show Groupings and Respetive Counts
+print('\nGrouped & Count by "emotion"')
+df_german_grouped.show(truncate=0)
+
+# Save Dataframe to CSV
+directory_path = 'data/spark_data_parts'
+df_german_exploded.coalesce(1).write.csv(directory_path, header=True, mode="overwrite")
+
+file_pattern = 'part-00000*.csv'
+file_path = glob.glob(directory_path + '/' + file_pattern)[0]
+
+shutil.move(file_path, './data/data_de.csv')
 \end{minted}

### 2023-07-18 paper/main.tex
Paper Updates

@@
 \usepackage[utf8]{inputenc}
+\usepackage[toc,page]{appendix}
 \usepackage{microtype}
@@
 \usepackage{minted}
-\setminted[]{linenos, samepage, numbersep=5pt, frame=lines}
+\setminted[]{linenos, numbersep=5pt, frame=lines}
 
@@
 
-\begin{wrapfigure}{r}{0.5\textwidth} %this figure will be at the right
+\begin{wrapfigure}{r}{0.4\textwidth} %this figure will be at the right
     \centering
-    \includegraphics[width=0.5\textwidth]{Plutchik-Wheel}
+    \includegraphics[width=0.35\textwidth]{Plutchik-Wheel}
     \caption{Plutchik Wheel of Emotions}
@@
 \begin{itemize}
-\item Joy: A feeling of happiness, contentment, or delight.
+\item Joy (Fun): A feeling of happiness, contentment, or delight.
 \item Sadness: A state of unhappiness, sorrow, or grief.
 \item Anger: An intense emotional response associated with feelings of displeasure, frustration, or hostility.
-\item Fear: An emotional response triggered by perceived threats or danger, often associated with anxiety or terror.
+\item Worry: An emotional response triggered by perceived threats or danger, often associated with anxiety or terror.
 \item Surprise: A sudden and unexpected reaction to something unusual or startling.
-\item Disgust: A strong aversion or revulsion towards something unpleasant, offensive, or repulsive.
+\item Disgust (Hate): A strong aversion or revulsion towards something unpleasant, offensive, or repulsive.
 \item Trust: A sense of reliance, confidence, or faith in someone or something.
@@
 It's important to note that emotions are complex and multifaceted, and different models may propose alternative categorizations or variations in the number of primary emotions.
-However for the scope of this research project the most common 7-8 emotions are used as described above.
+However for the scope of this research project the most common 8 emotions are used as described above.
+Both the English and German datasets being used in this research matched to the 8 listed above.
+Details on the linkage of the two datasets is described in the next section with an emphasis on the Table \ref{table:dataset_labels} detailed linkage.
 
 \subsection{Background and significance of Emotion Detection (ED) analysis in Machine Learning models}
-Emotion Detection (ED) analysis in Machine Learning models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific target\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions to name a few.
+Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions to just name a few.
 
 By accurately identifying and understanding emotions from text data, ML models can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.
+With the goal in mind of improving ML models this research paper came to be.
 
@@
 
-Due to the before mentioned this field of research provides significance in enabling machines to comprehend and respond appropriately to human emotions, contributing to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications that are growing rapidly.
+Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human emotions.
+This in turn contributes to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on.
 
 \subsection{ED challenges, including lack of comprehensive labeled datasets and language fragmentation}
-ED is growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes is much smaller.
+ED is growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller.
 The primary reason of the lack of data stems from the emotion detection requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created by humans.
@@
 \subsection{Motivation for the research project focusing on extending ED datasets through translation and evaluating the impact on prediction rates}
+
+This paper is significant because an entire industry of automated purported emotion reading (Text, Image, Video, and Sound) technologies is quickly emerging \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
+The market for emotion recognition software is forecast to reach at least \$3.8 billion by 2025 \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
+And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, and a multiple of others, it is fitting to work on research to solve issues.
+
 With the before mentioned issues regarding the quantity/lack and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED.
@@
 
-\begin{wrapfigure}{r}{0.4\textwidth} %this figure will be at the right
+\begin{wrapfigure}{R}{0.4\textwidth} %this figure will be at the right
     \centering
-    \includegraphics[width=0.4\textwidth]{Dataset-Extension}
+    \includegraphics[width=0.35\textwidth]{Dataset-Extension}
     \caption{Example of English dataset extension by translating from German.}
@@
 The first will be in English and the second in German.
-Each dataset will be translated in the other’s language for extending purposes.
-As an example figure \ref{fig:dataset-extension}, the English will be translated to German and the German to English by which the original datasets are extended.
+Each dataset will be translated into the other’s language for extending purposes.
+As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.
 
-Once the two datasets have been translated, we now extend the original language dataset with the translated once.
-In essence extending the datasets to create a larger one.
+The reason for selecting German was based on the author\footnote{Richard Hoehn} being fluent in English and German and hence being able to easily traverse the two languages while working on the code for dataset translations and formatting.
+
+Once the two datasets have been translated, we now extend the original language dataset with the translated one.By this approach we are in essence extending the original datasets to create a larger one wiht the foreign language data.
 With both extended datasets we can now train our ML models for English and German.
@@
 \subsubsection{PySpark Application Development for Translation \& Prediction}
-A program will be built in Jupiter notebooks that uses translations APIs to convert an English sentence to German and the same if a German sentence is fed to English.
-Both sentences are then processed in real-time on both the English and German models for the Emotion Detection and compared.
-The average of the prediction will be used for final prediction of the emotion that was processed.
+A program will be built in Jupiter Notebooks that uses translations APIs to convert an English sentence to German and the same if a German sentence is fed to English.
+Both sentences are then processed in real-time on both the English and German models for the Emotion Detection and will be compared to each other's output.
+The average of the prediction will be used for final prediction of the supplied sentence and the emotion that was detected.
+
+In order for such to be processed in real-time, the Jupyter Notebook will start a PySpark session that is accessed via the local server running that accepts RESTful API calls with details of the server in Appendix \ref{appendix:code-api-server}.
 
@@
     Boredom    &  179 & ---           & --- & NO  \\
-    Love       & 3842 & Vertrauen     & 130 & YES \\
+    Love       & 3842 & Vertrauen     & 316 & YES \\
     Relief     & 1526 & ---           & --- & NO  \\ 
     Fun        & 1776 & ---           & --- & NO  \\
-    Hate       & 1323 & Ekel          &   4 & YES \\
+    Hate       & 1323 & Ekel          &  29 & YES \\
     Neutral    & 8638 & Unklar        & 314 & YES \\
-    Anger      &  110 & Ärger         & 114 & YES \\
-    Happiness  & 5209 & Freude        &  58 & YES \\
-    Surprise   & 2187 & Überraschung  & 154 & YES \\
-    Sadness    & 5165 & Traurigkeit   & 101 & YES \\
-    Worry      & 8459 & Angst         &   6 & YES \\
-    Enthusiasm &  759 & Antizipation  & 413 & YES \\
+    Anger      &  110 & Ärger         & 226 & YES \\
+    Happiness  & 5209 & Freude        & 140 & YES \\
+    Surprise   & 2187 & Überraschung  & 369 & YES \\
+    Sadness    & 5165 & Traurigkeit   & 184 & YES \\
+    Worry      & 8459 & Angst         & 154 & YES \\
+    Enthusiasm &  759 & Antizipation  & 774 & YES \\
     Empty      &  827 & ---           & --- & NO  \\
@@
 
+Details on how we formatted the the two (English and German) datasets to each other can be viewed in the Appendix \ref{appendix:dataset_english} and \ref{appendix:dataset_german}
+
 \subsection{Development of a translation application for converting different languages into English and vice versa}
@@
 \begin{quote}
-\begin{minted}{python}
+\begin{minted}[fontsize=\small]{python}
 # Using Deep Translator to leveage Google Translate

### 2023-07-30 paper/appendix/dataset.tex
Paper Updates

@@
 # Rename Columns
-df_english_filtered = df_english_filtered.withColumnRenamed("sentiment", "emotion")
-df_english_filtered = df_english_filtered.withColumnRenamed("content", "sentence")
+df_english_filtered = df_english_filtered.withColumnRenamed("sentiment", "emotion_en")
+df_english_filtered = df_english_filtered.withColumnRenamed("content", "sentence_en")
 
 # Make sure the column order to the same for both german and english csv files
-df_english_filtered = df_english_filtered.select('sentence', 'emotion')
+df_english_filtered = df_english_filtered.select('sentence_en', 'emotion_en')
 
@@
 # Group By for Details & Count
-df_english_grouped = df_english_filtered.groupBy('emotion').count()
+df_english_grouped = df_english_filtered.groupBy('emotion_en').count()
 
@@
 shutil.move(file_path, './data/data_en.csv')
+
+Count (raw): 40000
+English Source Columns: ['tweet_id', 'sentiment', 'content']
+
+Count (filtered): 35692
+
+Grouped & Count by "emotion"
++----------+-----+
+|emotion_en|count|
++----------+-----+
+|love      |3842 |
+|hate      |1323 |
+|neutral   |8638 |
+|anger     |110  |
+|happiness |5209 |
+|surprise  |2187 |
+|sadness   |5165 |
+|worry     |8459 |
+|enthusiasm|759  |
++----------+-----+
 \end{minted}
@@
 # In order to "explode" the column to it's distinct rows
-df_german_exploded = df_german.select('*', explode('article_emotion').alias('emotion'))
+df_german_exploded = df_german.select('*', explode('article_emotion').alias('emotion_de'))
 
 # Rename Column
-df_german_exploded = df_german_exploded.withColumnRenamed("title", "sentence")
+df_german_exploded = df_german_exploded.withColumnRenamed("title", "sentence_de")
 
@@
 # Make sure the column order to the same for both german and english csv files
-df_german_exploded = df_german_exploded.select('sentence', 'emotion')
+df_german_exploded = df_german_exploded.select('sentence_de', 'emotion_de')
 
@@
 # Group By for Details & Count
-df_german_grouped = df_german_exploded.groupBy('emotion').count()
+df_german_grouped = df_german_exploded.groupBy('emotion_de').count()
 
@@
 shutil.move(file_path, './data/data_de.csv')
+
+Count (raw): 1970
+German Source Columns: ['article_emotion', 'article_id', 'article_stance', 'paragraphs', 'snippet', 'source', 'title']
+
+Count (filtered): 2568
+
+Grouped & Count by "emotion"
++------------+-----+
+|emotion_de  |count|
++------------+-----+
+|Vertrauen   |316  |
+|Freude      |140  |
+|Ärger       |226  |
+|Überraschung|369  |
+|Traurigkeit |184  |
+|Antizipation|774  |
+|Unklar      |314  |
+|Angst       |154  |
+|Ekel        |29   |
+|Keine       |62   |
++------------+-----+
 \end{minted}
+\clearpage
 
+\subsection{Details on Extending \& Translation}
+\label{appendix:dataset_translation}
+Details on translation and extending the datasets.
+\begin{minted}[fontsize=\small]{python}
+# **************************************
+# *** English <-> German Emotion Key ***
+# **************************************
+
+from pyspark.sql.functions import col, udf, split, explode, lit
+from pyspark.sql.types import IntegerType, StringType, StructType, StructField, StringType
+
+# Emotion Dictionary English <-> German
+# This key/value setup was previsoulsy created for linking purposes
+emotion_key = {
+    "boredom"    : "---",
+    "love"       : "Vertrauen",
+    "relief"     : "---",
+    "fun"        : "---",
+    "hate"       : "Ekel",
+    "neutral"    : "Unklar",
+    "anger"      : "Ärger",
+    "happiness"  : "Freude",
+    "surprise"   : "Überraschung", 
+    "sadness"    : "Traurigkeit", 
+    "worry"      : "Angst",
+    "enthusiasm" : "Antizipation", 
+    "empty "     : "---",
+    "---"        : "Keine",
+}
+
+# Create the schema for the DataFrame
+schema = StructType([
+    StructField("emotion_en", StringType(), nullable=False),
+    StructField("emotion_de", StringType(), nullable=False)
+])
+
+# Convert the dictionary to a list of tuples
+data = [(key, value) for key, value in emotion_key.items()]
+
+# Create the PySaprk DataFrame
+df_emotion_key = sc.createDataFrame(data, schema)
+
+df_emotion_key.show()
+
+# Extend the English and German onto the datasets
+df_german_exploded = df_german_exploded.join(df_emotion_key, on="emotion_de", how="left")
+df_english_filtered = df_english_filtered.join(df_emotion_key, on="emotion_en", how="left")
+
+# New sentence column
+df_german_exploded = df_german_exploded.withColumn("sentence_en", lit(None))
+df_english_filtered = df_english_filtered.withColumn("sentence_de", lit(None))
+
+df_german_exploded.show()
+df_english_filtered.show()
+
+# Using Deep Translator to proxy to Google Translate
+from deep_translator import GoogleTranslator
+
+# Define Tranlation Functions
+def translate_en_to_de(row):
+    translated = GoogleTranslator(source='en', target='de').translate(row["sentence_en"])
+    row["sentence_de"] = translated
+
+def translate_de_to_en(row):
+    translated = GoogleTranslator(source='de', target='en').translate(row["sentence_en"])
+    row["sentence_en"] = translated
+
+df_german_exploded.foreach(translate_de_to_en)
+
++----------+------------+
+|emotion_en|  emotion_de|
++----------+------------+
+|   boredom|         ---|
+|      love|   Vertrauen|
+|    relief|         ---|
+|       fun|         ---|
+|      hate|        Ekel|
+|   neutral|      Unklar|
+|     anger|       Ärger|
+| happiness|      Freude|
+|  surprise|Überraschung|
+|   sadness| Traurigkeit|
+|     worry|       Angst|
+|enthusiasm|Antizipation|
+|    empty |         ---|
+|       ---|       Keine|
++----------+------------+
+
++------------+--------------------+----------+----------+-----------+----------+
+|  emotion_de|         sentence_de|emotion_en|emotion_en|sentence_en|emotion_en|
++------------+--------------------+----------+----------+-----------+----------+
+|   Vertrauen|Adoptiert zu sein...|      love|      love|       null|      love|
+|      Freude|Österreichs Verfa...| happiness| happiness|       null| happiness|
+|      Freude|Visana gewinnt er...| happiness| happiness|       null| happiness|
+\end{minted}
 \clearpage

### 2023-07-30 paper/main.tex
Paper Updates

@@
 \section{Introduction}
+
+Emotion detection has gained significant interest in recent years as a means to understand human behavior and communication better.
+By analyzing text or speech, these detection processes can accurately identify and classify emotions expressed by individuals or groups.
+By employing a variety of ED processes, including natural language processing and machine learning, we can detect emotions with high precision.
+These ED systems find applications in diverse fields, such as sentiment analysis, customer feedback analysis, and mental health monitoring, contributing to a deeper understanding of human emotions and their impact on various aspects of life.
+
 \subsection{Background on Human Emotion Theoretical Framework}
 
-\begin{wrapfigure}{r}{0.4\textwidth} %this figure will be at the right
-    \centering
-    \includegraphics[width=0.35\textwidth]{Plutchik-Wheel}
-    \caption{Plutchik Wheel of Emotions}
-    \label{fig:plutchik-wheel}
-\end{wrapfigure}
-
 The concept of the eight main emotions can vary depending on the theoretical framework or model being considered.
-One well-known model is the Plutchik's Wheel of Emotions figure \ref{fig:plutchik-wheel}, which proposes eight primary emotions and their contrasting pairs.
-Here are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp}:
+One well-known model is the Plutchik's Wheel of Emotions Figure \ref{fig:plutchik-wheel}, which proposes eight primary emotions and their contrasting pairs.
+Below are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp}, a detailed reference of emotions used in this research can be found in Table \ref{table:dataset_labels} on page~\pageref{table:dataset_labels}.
 
@@
 \end{itemize}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.4\textwidth]{Plutchik-Wheel}
+    \caption{Plutchik Wheel of Emotions}
+    \label{fig:plutchik-wheel}
+\end{figure}
+
 It's important to note that emotions are complex and multifaceted, and different models may propose alternative categorizations or variations in the number of primary emotions.
 However for the scope of this research project the most common 8 emotions are used as described above.
-Both the English and German datasets being used in this research matched to the 8 listed above.
-Details on the linkage of the two datasets is described in the next section with an emphasis on the Table \ref{table:dataset_labels} detailed linkage.
+Both the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets being used in this research matched to the 8 listed above.
+Details on the linkage of the two datasets is described in Section \ref{sec:data-procurement-process} with an emphasis on the Table \ref{table:dataset_labels}'s detailed linkage.
 
@@
 By accurately identifying and understanding emotions from text data, ML models can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.
-With the goal in mind of improving ML models this research paper came to be.
+As an example, the primary goals of chatbots are entertainment, social contact, and novelty interaction, with a strong emphasis on productivity.
+In chatbots, there has been a tremendous requirement to simulate human-like characteristics and behavior.With the goal in mind of improving ML models this research paper came to be \cite{emotion-detection-literature-review}.
+These chatbots also improve customer engagement by offering friendliness, comfortness, flexibility, and efficient assistance.
+Chatbots should provide customers with more engaging responses, directly addressing their issues.
+Users perceive chatbots as companions rather than simple assistants.
+The majority of user requests are emotional rather than informative.
+Chatbots gained the capacity to respond emotionally to customers due to machine learning and sentiment analysis evolution\cite{chatbot-cognitive-awareness}.
 
@@
 \subsection{ED challenges, including lack of comprehensive labeled datasets and language fragmentation}
-ED is growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller.
+ED is growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data}.
 The primary reason of the lack of data stems from the emotion detection requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created by humans.
@@
 
-This paper is significant because an entire industry of automated purported emotion reading (Text, Image, Video, and Sound) technologies is quickly emerging \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
+This paper is significant because an entire industry of automated emotion reading (Text, Image, Video, and Sound) technologies is quickly emerging \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
 The market for emotion recognition software is forecast to reach at least \$3.8 billion by 2025 \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
-And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, and a multiple of others, it is fitting to work on research to solve issues.
+And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, customer service and a multiple of others, it is fitting to work on research to solve issues with prediction rates and lack of teaching / learning data.
 
-With the before mentioned issues regarding the quantity/lack and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED.
-The hope is that by use of translation that the predictability of ED can be improved by use of real- real-time language conversion and processing.
+With the before mentioned issues regarding the quantity (lack of) and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED.
+The hope is that by use of translation that the predictability of ED can be improved by use of real-time language conversion and processing.
 
@@
 
-\begin{wrapfigure}{R}{0.4\textwidth} %this figure will be at the right
+The procurement of two ED datasets is needed.
+The first will be in English and the second in German.
+Each dataset will be translated into the other’s language for extending purposes.
+As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.
+
+\begin{figure}[h]
     \centering
@@
     \label{fig:dataset-extension}
-\end{wrapfigure}
-
-The procurement of two ED datasets is needed.
-The first will be in English and the second in German.
-Each dataset will be translated into the other’s language for extending purposes.
-As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.
+\end{figure}
 
@@
 \subsection{Data procurement process, including obtaining ED datasets in English and German}
+\label{sec:data-procurement-process}
 The data procurement was relatively straight forward and in essence entailed using Google search terms for the Emotion Detection in English and German.
@@
 \subsection{Development of a translation application for converting different languages into English and vice versa}
-Translation of the emotion datasets was performed with Google Translator based on the Medium article\cite{Nidhaloff_how_to_translate_text_with_python} in which the author describes option fro simple API translations.
+Translation of the emotion datasets was performed with Google Translator based on the Medium article\cite{Nidhaloff_how_to_translate_text_with_python} in which the author describes option for simple API translations.
 As an example the following is a brief introduction to Google Translator's API:

### 2023-07-31 paper/appendix/dataset.tex
More Updates

--- 2023-07-18 paper/appendix/dataset.tex
@@
 The following details in this appendix are for clarification purposes.
-All examples in this appendix are created via PySpark using dataframes.
+All examples in this appendix are created via Panda and PySpark using data frames.
 
-\subsection{Using PySpark to Load Data}
-All example code was created using Jupyter Notebooks creating a PySpark session.
-The deails of setting up such session are below in the code example.
-All subsequent details are derived from the \texttt{sc}\footnote{sc = Spark Session}.
-\begin{minted}{python}
-# Setup Spark Session
-# Spark Session Name: Qualyfing_Exam
-from pyspark.sql import SparkSession
+\subsection{Translation Setup}
+All example code was created using Jupyter Notebooks.
+\begin{minted}[fontsize=\small]{python}
+# *********************
+# *** Setup Section ***
+# *********************
 
-# Get Imports Needed
-from pyspark.sql.functions import col, udf
+import pandas as pd
 
-# Get Datatypes needed for DataFrame manipulation
-from pyspark.sql.types import IntegerType, StringType
+# Emotion DataFrame
+# This was predtermined by review of the emotion from English to German
+emotion_key = {
+    "boredom": "---",
+    "love": "Vertrauen",
+    "relief": "---",
+    "fun": "---",
+    "hate": "Ekel",
+    "neutral": "Unklar",
+    "anger": "Ärger",
+    "happiness": "Freude",
+    "surprise": "Überraschung",
+    "sadness": "Traurigkeit",
+    "worry": "Angst",
+    "enthusiasm": "Antizipation",
+    "empty ": "---",
+    "---": "Keine"
+}
 
-# Setup Spark Session
-sc = SparkSession \
-        .builder \
-        .master("local[*]") \
-        .appName("Qualyfing_Exam") \
-        .getOrCreate()
+# Create a DataFrame from the emotion_key dictionary
+df_emotions = pd.DataFrame(emotion_key.items(), columns=['emotion_en', 'emotion_de'])
+
+# *******************
+# *** Translation ***
+# *******************
+from googletrans import Translator
+import time
+
+# This google API take a Sentence and converts to German
+def translate(sentence, dest_lang):
+    try:
+        translator = Translator()
+        translator.raise_Exception = True
+        translation = translator.translate(sentence, dest=dest_lang)
+        time.sleep(0.5)  # Add a 1-second delay
+        return translation.text
+    except Exception as e:
+        print(f"Translation Error: {e}")
+        return None
 \end{minted}
+\clearpage
 
 \subsection{Details on English Dataset}
-The English dataset based on tweets\footnote{Tweets from Twitter} are saved as a \texttt{csv}\footnote{csv = Comma Separated Values} file.
-The following are some simply details on the laod and processing of such file.
-\begin{minted}{python}
-# Read csv file to Saprk dataframe
-df_english = sc.read.csv("data/english.csv", header=True, inferSchema=True)
-# Group by a column and count the occurrences in df_english
-count_english = df_english.groupBy('sentiment').count()
+\label{appendix:dataset_english}
+The English dataset is based on tweets\footnote{Tweets from Twitter} that are saved as a \texttt{csv}\footnote{csv = Comma Separated Values} file.
+The following code was use to filter and process the file for use in this project.
+\begin{minted}[fontsize=\small]{python}
+# *********************************
+# *** English Data Preparations ***
+# *********************************
 
-# Show the result for df_english
-count_english.show()
-+----------+-----+
-| sentiment|count|
-+----------+-----+
-|   boredom|  179|
-|      love| 3842|
-|    relief| 1526|
-|       fun| 1776|
-|      hate| 1323|
-|   neutral| 8638|
-|     anger|  110|
-| happiness| 5209|
-|  surprise| 2187|
-|   sadness| 5165|
-|     worry| 8459|
-|enthusiasm|  759|
-|     empty|  827|
-+----------+-----+
+import pandas as pd
+from tqdm import tqdm
 
+# Load English CSV File into a Dataframe
+df_en = pd.read_csv("data/english.csv")
+print(f'Count (raw): {len(df_en)}')
+
+# Print Columns
+print(f'English Source Columns: {df_en.columns}\n')
+
+# Filter only labels that we matched with the German counter part JSON File
+filter_values = ["love", "hate", "neutral", "anger", "happiness", "surprise", "sadness", "worry", "enthusiasm"]
+df_en = df_en[df_en["sentiment"].isin(filter_values)]
+
+# Find Empty Sentence Values
+df_en = df_en[df_en["content"].notnull() & (df_en["content"].str.strip() != '')]
+
+# Remove unnecessary column
+df_en = df_en.drop("tweet_id", axis=1)
+
+# Rename Columns
+df_en = df_en.rename(columns = {'sentiment': 'emotion_en', 'content': 'sentence_en'})
+
+# Merge German Emotions onto English
+df_en = pd.merge(df_en, df_emotions, on='emotion_en', how='left')
+
+# Add German Sentence Column
+df_en["sentence_de"] = ""
+
+# Randomly select 1500 rows
+df_en = df_en.sample(n=1500, random_state=2023)
+
+df_en.to_csv('./data/pd_en.csv', index=False)
+
+# Iterate over the rows with tqdm to show the progress
+for index, row in tqdm(df_en.iterrows(), total=df_en.shape[0]):
+    # Call Translation
+    sentence = translate(row["sentence_en"], 'de')
+    # Save Sentence on Column
+    df_en.at[index, 'sentence_de'] = sentence
+
+df_en.to_csv('./data/pd_en_translated.csv', index=False)
 \end{minted}
+\clearpage
 
 \subsection{Details on German Dataset}
+\label{appendix:dataset_german}
 The German dataset was downloaded from the ETH's\footnote{ETH (German: Eidgenössische Technische Hochschule) or Swiss Federal Institute of Technology} servers via "link" as a \texttt{JSON} file.
-\begin{minted}{python}
-# Read JSON file to Saprk dataframe
-df_german = sc.read.json("data/german.json")
+Since the the file was \texttt{JSON} we will need to filter and explode the file based on teh emotions that are attached to it.
+\begin{minted}[fontsize=\small]{python}
+# *********************************
+# *** German Data Preparations ***
+# *********************************
 
-# Group by a column and count the occurrences in df_german
-count_german = df_german.groupBy('article_emotion').count()
+import pandas as pd
+from tqdm import tqdm
 
-# Show the result for df_german
-count_german.show(truncate=0)
-+----------------------------------+-----+
-|article_emotion                   |count|
-+----------------------------------+-----+
-|[Ärger, Antizipation, Traurigkeit]|1    |
-|[Überraschung, Vertrauen]         |14   |
-|[Überraschung, Freude]            |15   |
-|[Ärger, Vertrauen]                |7    |
-|[Ekel, Traurigkeit]               |4    |
-|[Keine, Antizipation]             |5    |
-|[Vertrauen]                       |130  |
-|[Freude]                          |58   |
-|[Ärger, Antizipation]             |31   |
-|[Ärger]                           |114  |
-|[Überraschung]                    |154  |
-|[Antizipation, Traurigkeit]       |15   |
-|[Antizipation, Freude, Vertrauen] |2    |
-|[Vertrauen, Angst]                |5    |
-|[Traurigkeit]                     |101  |
-|[Antizipation]                    |413  |
-|[Traurigkeit, Vertrauen]          |7    |
-|[Antizipation, Vertrauen]         |104  |
-|[Ärger, Traurigkeit, Angst]       |1    |
-|[Unklar]                          |314  |
-+----------------------------------+-----+
+# Load German JSON File into a DataFrame
+df_de = pd.read_json("data/german.json", lines=True)
+print(f'Count (raw): {len(df_de)}')
 
+# Print Columns
+print(f'German Source Columns: {df_de.columns}\n')
+
+# Explode the "article_emotion" column to distinct rows
+df_de = df_de.explode('article_emotion')
+
+# Filter only labels that we matched with the German counter part JSON File
+filter_values = ["Vertrauen","Ekel", "Unklar", "Ärger", "Freude", "Überraschung", "Traurigkeit", "Angst", "Antizipation"]
+df_de = df_de[df_de["article_emotion"].isin(filter_values)]
+
+# Find Empty Sentence Values
+df_de = df_de[df_de["title"].notnull() & (df_de["title"].str.strip() != '')]
+
+# Remove unnecessary columns
+df_de = df_de.drop("article_id", axis=1)
+df_de = df_de.drop("article_stance", axis=1)
+df_de = df_de.drop("paragraphs", axis=1)
+df_de = df_de.drop("source", axis=1)
+df_de = df_de.drop("snippet", axis=1)
+
+# Rename Columns
+df_de = df_de.rename(columns={'title': 'sentence_de'})
+df_de = df_de.rename(columns={'article_emotion': 'emotion_de'})
+
+# Merge Englsih Emotions onto German
+df_de = pd.merge(df_de, df_emotions, on='emotion_de', how='left')
+
+# Add English Sentence Column
+df_de["sentence_en"] = ""
+
+# Randomly select 1500 rows
+df_de = df_de.sample(n=1500, random_state=2023)
+
+df_de.to_csv('./data/pd_de.csv', index=False)
+
+# Iterate over the rows with tqdm to show the progress
+for index, row in tqdm(df_de.iterrows(), total=df_de.shape[0]):
+    # Call Translation
+    sentence = translate(row["sentence_de"], 'en')
+    # Save Sentence on Column
+    df_de.at[index, 'sentence_en'] = sentence
+
+df_de.to_csv('./data/pd_de_translated.csv', index=False)
 \end{minted}
-
 \clearpage

### 2023-07-31 paper/main.tex
More Updates

@@
 \setlength{\parskip}{1em}
+
+\usepackage{array}
+\newcolumntype{L}{>{\centering\arraybackslash}m{1.2in}}
 
@@
 \begin{abstract}
-This Qualifying Exam\footnote{PhD students of MTSU Computational and Data  Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages.
+This Qualifying Exam\footnote{PhD students of MTSU Computational and Data  Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages (English and German).
 The research tries to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis.
 
-By extending an original language dataset with a translated dataset from another language and utilizing parallel processing with PySpark, this research project aims to achieve higher prediction rates in ED.
+By extending an original language dataset with a translated dataset, in the case of this paper English and German.
+And utilizing parallel processing with PySpark, this research project aims to achieve higher prediction rates in ED.
 The results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper demonstrate the effectiveness of such an approach and how the real-time translation technique can enhance emotion detection capabilities.
@@
 
-Emotion detection has gained significant interest in recent years as a means to understand human behavior and communication better.
-By analyzing text or speech, these detection processes can accurately identify and classify emotions expressed by individuals or groups.
-By employing a variety of ED processes, including natural language processing and machine learning, we can detect emotions with high precision.
-These ED systems find applications in diverse fields, such as sentiment analysis, customer feedback analysis, and mental health monitoring, contributing to a deeper understanding of human emotions and their impact on various aspects of life.
+Emotion detection has gained significant interest in recent years as a means to understand human behavior and improve communication effectiveness.
+By analyzing text or speech, these detection processes can accurately identify and classify emotions expressed by individuals or groups, often times in real-time.
 
-\subsection{Background on Human Emotion Theoretical Framework}
+By employing a variety of ED processes, including natural language processing and machine learning, applications can detect emotions with high precision.
+These ED systems find applications in diverse fields, such as sentiment analysis, customer feedback analysis, voting and news stances, and mental health monitoring, which all aim at contributing to a deeper understanding of human emotions and their impact on various aspects of life.
 
-The concept of the eight main emotions can vary depending on the theoretical framework or model being considered.
-One well-known model is the Plutchik's Wheel of Emotions Figure \ref{fig:plutchik-wheel}, which proposes eight primary emotions and their contrasting pairs.
+\subsection{Human Emotion as a Theoretical Framework}
+
+The concept and number of emotions can vary depending on the theoretical framework or model being considered.
+One well-known model is the Plutchik's Wheel of Emotions Figure \ref{fig:plutchik-wheel}, which proposes eight (8) primary emotions and their contrasting pairs.
+In this paper we will be focusing on the basic eight emotions and not on their sub-emotions.
+
 Below are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp}, a detailed reference of emotions used in this research can be found in Table \ref{table:dataset_labels} on page~\pageref{table:dataset_labels}.
@@
 \begin{itemize}
-\item Joy (Fun): A feeling of happiness, contentment, or delight.
+\item Happiness: A feeling of Joy (Fun), contentment, or delight.
 \item Sadness: A state of unhappiness, sorrow, or grief.
@@
 \item Surprise: A sudden and unexpected reaction to something unusual or startling.
-\item Disgust (Hate): A strong aversion or revulsion towards something unpleasant, offensive, or repulsive.
-\item Trust: A sense of reliance, confidence, or faith in someone or something.
-\item Anticipation: A feeling of excited expectation or eagerness towards future events.
+\item Disgust and/or Hate: A strong aversion or revulsion towards something unpleasant, offensive, or repulsive.
+\item Trust and/or Love: A sense of reliance, confidence, or faith in someone or something.
+\item Enthusiasm: A feeling of excited expectation or eagerness towards future events.
 \end{itemize}
@@
 It's important to note that emotions are complex and multifaceted, and different models may propose alternative categorizations or variations in the number of primary emotions.
-However for the scope of this research project the most common 8 emotions are used as described above.
-Both the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets being used in this research matched to the 8 listed above.
-Details on the linkage of the two datasets is described in Section \ref{sec:data-procurement-process} with an emphasis on the Table \ref{table:dataset_labels}'s detailed linkage.
+However for the scope of this research project the most common eight (8) emotions are used as described above.
 
-\subsection{Background and significance of Emotion Detection (ED) analysis in Machine Learning models}
-Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions to just name a few.
+By aligning the emotional labels across the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets being used in this research all eight listed above were used.
+Details on the linkage of the two datasets is described in Section \ref{sec:data-procurement-process} with an emphasis on the Table \ref{table:dataset_labels}'s cross reference on page~\pageref{table:dataset_labels}.
 
-By accurately identifying and understanding emotions from text data, ML models can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.
-As an example, the primary goals of chatbots are entertainment, social contact, and novelty interaction, with a strong emphasis on productivity.
-In chatbots, there has been a tremendous requirement to simulate human-like characteristics and behavior.With the goal in mind of improving ML models this research paper came to be \cite{emotion-detection-literature-review}.
-These chatbots also improve customer engagement by offering friendliness, comfortness, flexibility, and efficient assistance.
+\subsection{Significance of Emotion Detection (ED) in Machine Learning}
+Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions (chatbots\cite{chatbot-cognitive-awareness}) to just name a few.
+
+By accurately identifying and understanding emotions from text data, ML applications can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.
+As an example, the primary goals of chatbots are entertainment, social contact, and novelty interaction, with a strong emphasis on productivity, scale-ability and cost reductions.
+In chatbots, there has been a immense requirement to simulate human-like characteristics and behavior during human-machine interactions\cite{emotion-detection-literature-review}.
+Per Kusal et al.
+\cite{chatbot-cognitive-awareness} these chatbots also improve customer engagement by offering friendliness, comfortness, flexibility, and efficient assistance.
 Chatbots should provide customers with more engaging responses, directly addressing their issues.
-Users perceive chatbots as companions rather than simple assistants.
-The majority of user requests are emotional rather than informative.
+The goal of deploying ED applications is for users perceive chatbots as companions rather than simple assistants.
+Per Adikari et al.
+the majority of user requests are emotional rather than informative.
 Chatbots gained the capacity to respond emotionally to customers due to machine learning and sentiment analysis evolution\cite{chatbot-cognitive-awareness}.
@@
 
-Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human emotions.
-This in turn contributes to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on.
+Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions.
+This in turn contributes to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions}.
 
-\subsection{ED challenges, including lack of comprehensive labeled datasets and language fragmentation}
-ED is growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data}.
-The primary reason of the lack of data stems from the emotion detection requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created by humans.
+\subsection{ED challenges: Lack of Comprehensive Labeled Datasets and Language Fragmentation}
+ED is an immensely growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}.
+The primary reason of the lack of data stems from the emotion detection data requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created and labeled by humans.
 To compound on the issue of the lack of data; the many datasets that are available are in multiple languages.
+A significant amount of time took place to find the German labeled
 
@@
 \subsection{Objectives \& Scope of the Research}
-The project is comprised of three main points / goals which are listed in following sub-sections with details in each point as follows:
 
-\subsubsection{Data Collection / Translation / ML Model Training}
+The research project aims at answering the following research questions:
 
-The procurement of two ED datasets is needed.
-The first will be in English and the second in German.
-Each dataset will be translated into the other’s language for extending purposes.
-As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.35\textwidth]{Dataset-Extension}
-    \caption{Example of English dataset extension by translating from German.}
-    \label{fig:dataset-extension}
-\end{figure}
-
-The reason for selecting German was based on the author\footnote{Richard Hoehn} being fluent in English and German and hence being able to easily traverse the two languages while working on the code for dataset translations and formatting.
-
-Once the two datasets have been translated, we now extend the original language dataset with the translated one.By this approach we are in essence extending the original datasets to create a larger one wiht the foreign language data.
-With both extended datasets we can now train our ML models for English and German.
-
-\subsubsection{PySpark Application Development for Translation \& Prediction}
-A program will be built in Jupiter Notebooks that uses translations APIs to convert an English sentence to German and the same if a German sentence is fed to English.
-Both sentences are then processed in real-time on both the English and German models for the Emotion Detection and will be compared to each other's output.
-The average of the prediction will be used for final prediction of the supplied sentence and the emotion that was detected.
-
-In order for such to be processed in real-time, the Jupyter Notebook will start a PySpark session that is accessed via the local server running that accepts RESTful API calls with details of the server in Appendix \ref{appendix:code-api-server}.
-
-\subsubsection{Written \& Oral Presentation of Research Finding}
-In conclusion of the project a research paper and oral presentation is required to pass MTSU Computation Science PhD Qualifying Exam.
-Such paper is being presented with this paper and the oral presentation will be in Microsoft Power Point format.
-
-The listing the datasets used, code for the application, and details of the research project process in this research paper is openly available and hosted on a GitHub repository, allowing for easy access, reproducible, and collaborative engagement with the research findings.
-The repository is open and available under: \texttt{https://github.com/richardhoehn/mtsu.coms.qualifying-exam}\cite{Hoehn_Improving_Emotion_Detection_2023}
+\begin{itemize}
+\item Can by translating English data to German and extending an original German dataset increase the predictability of the ED model?
+\item Can by translating German data to English and extending an original English dataset increase the predictability of the ED model?
+\item Can by translating in real-time an input to multiple languages improve the predictability based on the combined output of two model's results.
+\end{itemize}
 
 \section{Literature Review}
+To ensure comprehensive coverage of the latest developments and research in ED for this research paper, consideration and review was only given to incorporating those research papers and online news articles that were published after the year 2018, thereby providing an up-to-date perspective on the subject matter.
 
-\subsection{Emotion Detection Importance, Data Limitations, Diverse Populations}
+\subsection{Emotion Detection Importance and Diverse Populations}
 Emotion Detection (ED) holds significant importance in various domains, including psychology, social sciences, and human-computer interaction.
 It enables the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states.
-However, the field faces limitations due to the scarcity of comprehensive labeled datasets.
+
+Furthermore, emotions can differ across age groups, genders, cultures, and languages.
+Including data from diverse populations helps in developing inclusive and culturally sensitive ED models.
+It ensures that the models are not biased towards specific groups and can accurately detect emotions in a wide range of individuals.
+
+\subsection{Lack of Labeled Datasets and the Fragmentation caused by Different Languages}
+
 The lack of diverse and large-scale data hinders the development and training of accurate ED models, limiting their effectiveness and general ability across different contexts and populations.
@@
 In recent research, efforts have been made to address these limitations by exploring new techniques for emotion detection.
-For instance, a study by Shalini Kapoor, Tarun Kumar \cite{KAPOOR2023120882} proposes analyzing the entire spectrum of the language source to predict emotional changes.
-
-Overcoming the scarcity of labeled datasets and embracing innovative approaches like the before mentioned study by Shalini Kapoor, Tarun Kumar \cite{KAPOOR2023120882} and their analysis can contribute to the advancement of ED research.
+For instance, a study by Kapoor\cite{KAPOOR2023120882} et al.
+proposes analyzing the entire spectrum of the language source to predict emotional changes.
+This provided some help, but quickly established itself as time consuming and need of expertise in the subject matters.
+Overcoming the scarcity of labeled datasets and embracing innovative approaches like the before mentioned study by Kapoor\cite{KAPOOR2023120882} et al.
+and their analysis can contribute to the advancement of ED research.
 Such advancements hold the potential to enhance the accuracy and applicability of emotion detection models, enabling a deeper understanding of human emotions and more effective responses in various domains.
 
-Emotions can differ across age groups, genders, cultures, and languages.
-Including data from diverse populations helps in developing inclusive and culturally sensitive ED models.
-It ensures that the models are not biased towards specific groups and can accurately detect emotions in a wide range of individuals.
+The paper by Mascarell\cite{mascarell-etal-2021-stance} et al.
+states that unlike images language operate very differently, therefore the lack of labeled datasets poses a significant challenge in text-based Emotion Detection (ED) analysis as it limits the availability of reliable training data for ML models.
+This scarcity hampers the model's ability to learn and generalize emotions effectively.
 
-\subsection{Challenges by the lack of Labeled Datasets and the Fragmentation caused by Different Languages}
-Unlike images language operate very differently, therefore the lack of labeled datasets poses a significant challenge in text-based Emotion Detection (ED) analysis as it limits the availability of reliable training data for ML models.
-This scarcity hampers the model's ability to learn and generalize emotions effectively.
-Additionally, the fragmentation caused by different languages further exacerbates the issue, as it reduces the size and diversity of data available for training, resulting in limited cross-lingual generalization and potentially biased models.
+Additionally, Kusal et al.
+further stipulates \cite{kusal} that the fragmentation caused by different languages further exacerbates the issue, as it reduces the size and diversity of data available for training, resulting in limited cross-lingual generalization and potentially biased models.
 Overcoming these challenges is crucial for advancing ED research and improving the accuracy and applicability of emotion detection models across various languages and cultures.
@@
 
-\subsection{Data procurement process, including obtaining ED datasets in English and German}
+\subsection{Data procurement process of ED datasets in English and German}
 \label{sec:data-procurement-process}
 The data procurement was relatively straight forward and in essence entailed using Google search terms for the Emotion Detection in English and German.
-Some of the main German data was obtained from the dataset built by ETH's Emotion and Stance Detection for German Text \cite{mascarell-etal-2021-stance} which also hosts a good website with their finds \footnote{https://mtc.ethz.ch/research/natural-language-processing/emotion-stance.html}.
+Some of the main German data was obtained from the dataset built by ETH's Emotion and Stance Detection for German Text \cite{mascarell-etal-2021-stance} which also hosts a good website with their findings \footnote{https://mtc.ethz.ch/research/natural-language-processing/emotion-stance.html}.
 The English dataset was downloaded from Kaggle \footnote{https://www.kaggle.com/datasets/pashupatigupta/emotion-detection-from-text} based on Tweets collected in 2021.
@@
 
+The procurement of two ED datasets is needed.
+The first will be in English and the second in German.
+Each dataset will be translated into the other’s language for extending purposes.
+As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.35\textwidth]{Dataset-Extension}
+    \caption{Example of English dataset extension by translating from German.}
+    \label{fig:dataset-extension}
+\end{figure}
+
+The reason for selecting German was based on the author\footnote{Richard Hoehn} being fluent in English and German and hence being able to easily traverse the two languages while working on the code for dataset translations and formatting.
+
+Once the two datasets have been translated, we now extend the original language dataset with the translated one.By this approach we are in essence extending the original datasets to create a larger one with the foreign language data.
+With both extended datasets we can now train our ML models for English and German.
+
 \subsection{Development of a translation application for converting different languages into English and vice versa}
@@
 
-\subsection{Building and training multiple ML models using the extended datasets and parallel processing approach with PySpark}
-\ldots
+Based on the above example the CSV and JSON datasets were individually loaded, parsed, and translated to their respective counter-part languages.
+Details of the application can be found in the Appendix \ref{appendix:dataset_english} for English and Appendix \ref{appendix:dataset_german} for the German.
+We decided to only use 1,500 row for each language so since the German (DE\footnote{DE = German}) dataset was not much larger than that.
+This meant that the English (EN\footnote{EN = Enlgish}) datasets was randomly paired down to 1,500 to match that of the German.
+
+Each translated file now contains four (4) columns that are labeled as in Tables \ref{table:pd_de_translated} and \ref{table:pd_en_translated} with the exact same column naming.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | L | L | L | L | }
+    \hline
+    \multicolumn{4}{|c|}{File: \texttt{pd\textunderscore de\textunderscore translated.csv}} \\
+    \hline
+    sentence\textunderscore de &
+    emotion\textunderscore de & 
+    sentence\textunderscore en & 
+    emotion\textunderscore en \\
+    \hline
+    Original Sentence (feature) &
+    Original Emotion (label) &
+    Sentence Translated to English & 
+    Emotion set by Cross-Reference from German \\
+
+    \hline
+\end{tabular}
+\caption{German CSV file structure after translation to English}
+\label{table:pd_de_translated}
+\end{table}
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | L | L | L | L | }
+    \hline
+    \multicolumn{4}{|c|}{File: \texttt{pd\textunderscore en\textunderscore translated.csv}} \\
+    \hline
+    sentence\textunderscore en &
+    emotion\textunderscore en & 
+    sentence\textunderscore de & 
+    emotion\textunderscore de \\
+    \hline
+    Original Sentence (feature) &
+    Original Emotion (label) &
+    Sentence Translated to German & 
+    Emotion set by Cross-Reference from English \\
+
+    \hline
+\end{tabular}
+\caption{English CSV File structure after translation to German}
+\label{table:pd_en_translated}
+\end{table}
+
+\subsection{Training Multiple ML models using the Extended Datasets with PySpark}
+
+ Both sentences are then processed in real-time on both the English and German models for the Emotion Detection and will be compared to each other's output.
+The average of the prediction will be used for final prediction of the supplied sentence and the emotion that was detected.
+
+ \begin{table}[h!]
+\centering
+\begin{tabular}{ | L | L | L | L | }
+    \hline
+    \multicolumn{4}{|c|}{ML Models } \\
+    \hline
+    Model & 
+    Data &
+    Rows for Training & 
+    Rows for Testing  \\
+    \hline
+    A &
+    English Original &
+    1200 & 
+    300 (20\%) \\
+    \hline
+    B &
+    English Extended By German &
+    2700 & 
+    300 same as Model "A" \\
+    \hline
+    C &
+    German Original &
+    1200 & 
+    300 (20\%) \\
+    \hline
+    D &
+    German Extended By English &
+    2700 & 
+    300 same as Model "C" \\
+
+    \hline
+\end{tabular}
+\caption{ML Models and their respective settings}
+\label{table:ml_models}
+\end{table}
+
+In order for such to be processed in real-time, the Jupyter Notebook will start a PySpark session that is accessed via the local server running that accepts RESTful API calls with details of the server in Appendix \ref{appendix:code-api-server}.
 
@@
 Metrics like accuracy, precision, recall, F1 score, or area under the receiver operating characteristic curve (AUC-ROC) are commonly used to measure the models' performance.
+
+\subsubsection{Written \& Oral Presentation of Research Finding}
+In conclusion of the project a research paper and oral presentation is required to pass MTSU Computation Science PhD Qualifying Exam.
+Such paper is being presented with this paper and the oral presentation will be in Microsoft Power Point format.
+
+The listing the datasets used, code for the application, and details of the research project process in this research paper is openly available and hosted on a GitHub repository, allowing for easy access, reproducible, and collaborative engagement with the research findings.
+The repository is open and available under: \texttt{https://github.com/richardhoehn/mtsu.coms.qualifying-exam}\cite{Hoehn_Improving_Emotion_Detection_2023}
 

### 2023-08-04 paper/main.tex
Paper Updates

@@
 \tableofcontents
+
 \clearpage
-
 \section{Introduction}
@@
 
+\clearpage
 \section{Literature Review}
@@
 It enables the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states.
+In Chowanda et.
+al.
+paper\cite{CHOWANDA-2021821} they believe that "Emotions hold a paramount role in the conversation, as it expresses context to the conversation."
 
@@
 
+\clearpage
 \section{Methodology}
@@
 
-Each translated file now contains four (4) columns that are labeled as in Tables \ref{table:pd_de_translated} and \ref{table:pd_en_translated} with the exact same column naming.
+Each translated file now contains four (4) columns that are labeled in Tables \ref{table:pd_de_translated} and \ref{table:pd_en_translated} with the exact same column naming.
 
@@
 
- Both sentences are then processed in real-time on both the English and German models for the Emotion Detection and will be compared to each other's output.
-The average of the prediction will be used for final prediction of the supplied sentence and the emotion that was detected.
+ Using the datasets from Tables \ref{table:pd_de_translated}, \ref{table:pd_en_translated}, and \ref{table:ml_models} that list an approximate $\sim$1,275 and $\sim$2,775 features for training.
+The algorithms tested in this research project ranged from the traditional machine learning to deep learning techniques, they are Na\"ive Bayes, Generalised Linear Model (GLM), and Random Forest Classifier models.
+Each of these algorithms used and their results are listed in the Results and Analysis Section \ref{sec:results-and-analysis} below.
+In Chowanda et al.
+paper\cite{CHOWANDA-2021821} on Text-based Emotion Techniques multiple algorithms were proposed, this research focused on just three out of their six since the others had lower prediction rates based on Chowanda et al.
+paper\cite{CHOWANDA-2021821} paper.
 
@@
     Data &
-    Rows for Training & 
-    Rows for Testing  \\
+    Rows for Training (85\%) & 
+    Rows for Testing (15\%)  \\
     \hline
@@
     English Original &
-    1200 & 
-    300 (20\%) \\
+    1,275 & 
+    225 \\
     \hline
@@
     English Extended By German &
-    2700 & 
-    300 same as Model "A" \\
+    2,775 & 
+    225 same as Model "A" \\
     \hline
@@
     German Original &
-    1200 & 
-    300 (20\%) \\
+    1,275 & 
+    225 \\
     \hline
@@
     German Extended By English &
-    2700 & 
-    300 same as Model "C" \\
+    2,775 & 
+    225 same as Model "C" \\
 
@@
 \subsection{Evaluation Metrics and methodologies for measuring the performance of the ML models}
-The evaluation of ML models trained on text datasets is a critical step in assessing their performance and reliability.
-Our research project is no different and the following assumptions are used to determine the effectiveness of the methodology of extending our datasets:
+The evaluation of ML models trained on the English, German and their respective extension datasets is a critical step in assessing their performance and reliability.
+We chose to use the MulticlassClassificationEvaluator\footnote{URL: https://spark.apache.org/docs/latest/api/python/reference/api/pyspark.ml.evaluation.MulticlassClassificationEvaluator.html} because is a well known tool in the field of natural language processing (NLP) research, specifically in multi class classification tasks such as Emotion Detection (ED).
+Generally this evaluator is used to evaluate the performance of machine learning (ML) models, using transformer-based architectures like BERT, RoBERTa, or DistilBERT\cite{Sanh2019-DistilBERT-AD}, with DistilBERT\cite{Sanh2019-DistilBERT-AD} being the one use in this research.
+
+The MulticlassClassificationEvaluator provides the essential evaluation metrics, including accuracy, precision, recall, and F1 score, for ED classifications.
+These metrics provide valuable insights into a model's ability to correctly classify emotions across multiple classes and language identification.
+Our research project is no different and the following that the assumptions are used to determine the effectiveness of the methodology of extending our datasets:
 \begin{itemize}
+
 \item \textbf{Accuracy:} Accuracy measures the overall correctness of the emotion predictions made by the model.
@@
 However, accuracy may not provide a complete picture, especially if the dataset is imbalanced.
+
 \item \textbf{Precision, Recall, and F1 Score:} Precision represents the proportion of correctly predicted emotions among all instances predicted as a specific emotion.
 Recall measures the proportion of correctly predicted emotions among all instances that truly belong to a specific emotion.
-F1 score combines precision and recall into a single metric, providing a balanced evaluation.
-\item \textbf{Cross-Validation:} Cross-validation techniques, such as k-fold or stratified cross-validation, divide the dataset into multiple subsets for training and testing.
-This helps evaluate the model's performance across different data samples, mitigating over-fitting and providing a more robust estimation of its generalization ability.
+F1 Score ($F_{1}=\frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}$) combines precision and recall into a single metric, providing a balanced evaluation.
+
 \item \textbf{Error Analysis:} Performing error analysis on the model's predictions helps identify specific patterns or instances where the model might struggle or make frequent mistakes.
@@
 
-\subsubsection{Written \& Oral Presentation of Research Finding}
+\subsection{Written \& Oral Presentation of Research Finding}
 In conclusion of the project a research paper and oral presentation is required to pass MTSU Computation Science PhD Qualifying Exam.
-Such paper is being presented with this paper and the oral presentation will be in Microsoft Power Point format.
-
 The listing the datasets used, code for the application, and details of the research project process in this research paper is openly available and hosted on a GitHub repository, allowing for easy access, reproducible, and collaborative engagement with the research findings.
@@
 
+\clearpage
 \section{Results and Analysis}
+\label{sec:results-and-analysis}
+The following section lists the prediction rates of training and testing of the original and extended datasets both in English and German.
+Overall results seem to indicate that there was no and even negative results based on adding the extended data for Emotion Detection.
+In the below sub sections details on the results are listed in prose, tables, and formauls.
 
-\subsection{Presentation of the prediction results obtained from the extended datasets}
+\subsection{Prediction Results of Original \& Extended Datasets}
 
@@
 
-\subsection{Comparison of Prediction Result of the Original and Extended Language (translated) Datasets}
+\subsection{Comparison of Prediction Rates of Original and Extended Language Datasets}
 \ldots
@@
 
+\subsection{Conclusion and Future Work}
+\ldots
+
+\clearpage
 \section{Demonstration and Application}

### 2023-08-08 paper/main.tex
Paper & Code Updates

@@
 \usepackage{multirow}
+\usepackage{tocloft}
 
@@
 \newcolumntype{L}{>{\centering\arraybackslash}m{1.2in}}
+\newcolumntype{M}{>{\centering\arraybackslash}m{0.85in}}
 
@@
 \begin{abstract}
-This Qualifying Exam\footnote{PhD students of MTSU Computational and Data  Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages (English and German).
-The research tries to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis.
+This Qualifying Exam\footnote{PhD students of MTSU Computational and Data Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages specifically  English and German.
+The research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis research and projects.
 
-By extending an original language dataset with a translated dataset, in the case of this paper English and German.
-And utilizing parallel processing with PySpark, this research project aims to achieve higher prediction rates in ED.
-The results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper demonstrate the effectiveness of such an approach and how the real-time translation technique can enhance emotion detection capabilities.
+By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) the hope is that better prediction rates can be achieved in ED analysis applications.
+Furthermore, translating English to German to extend the German dataset and training another ML model in parallel using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates by dual processing a sentence at the same time.
+
+In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, translated, and two ML models built that can be accessed via an API to get a prediction on an input (POST or GET) sentence both in German or English.
 \end{abstract}
@@
 
-\subsection{ED challenges: Lack of Comprehensive Labeled Datasets and Language Fragmentation}
+Lastly, in Chen et al.\cite{research-emotion-recognition-for-online-learning} paper based on emotion detection for online learning, it states that with emotion recognition research, we need to focus on accuracy and real-time performance in order to apply emotional recognition based on physiological signals to solve practical problems; they key part of their claim is that in order for ED to be practical real-time methods needs to be used.
+
+\subsection{ED challenges: Lack of Labeled Datasets and Language Fragmentation}
 ED is an immensely growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}.
@@
 \section{Literature Review}
-To ensure comprehensive coverage of the latest developments and research in ED for this research paper, consideration and review was only given to incorporating those research papers and online news articles that were published after the year 2018, thereby providing an up-to-date perspective on the subject matter.
+To ensure comprehensive coverage of the latest developments and research in emotion detection and analysis for this research paper, consideration and review was only given to incorporating research papers and online news articles that were published after the year 2015, thereby providing an up-to-date perspective on the subject matter.
+By implementing the 2015 boundary, our methodology aims to provide a more up-to-date outlook on the subject matter at hand and what areas the ED field in moving in in respect to ML applications and processes.
 
@@
 
-\subsection{Lack of Labeled Datasets and the Fragmentation caused by Different Languages}
+\subsection{Lack of Labeled Datasets and Fragmentation caused by Different Languages}
 
@@
 
-\subsection{Training Multiple ML models using the Extended Datasets with PySpark}
+\subsection{Training Multiple ML models with PySpark}
 
  Using the datasets from Tables \ref{table:pd_de_translated}, \ref{table:pd_en_translated}, and \ref{table:ml_models} that list an approximate $\sim$1,275 and $\sim$2,775 features for training.
-The algorithms tested in this research project ranged from the traditional machine learning to deep learning techniques, they are Na\"ive Bayes, Generalised Linear Model (GLM), and Random Forest Classifier models.
+The algorithms tested in this research project ranged from the traditional machine learning to deep learning techniques, they are:
+ \begin{itemize}
+
+\item \textbf{Na\"ive Bayes:} Per Pujan's paper\cite{naive-bayes-model} on the Na\"ive Bayes model, it classifies and/or assigns the most likely class to a given example based on its feature vector, simplifying the task greatly by assuming that the features are independent given a class.
+
+\item \textbf{Generalised Linear Model (GLM):} Per Mueller's paper on GLM\cite{glm-model} the Generalized linear models extend the concept of the well understood linear regression model.
+
+\item \textbf{Random Forest Classifier:} Per Schonlau's paper\cite{random-forest-model} on the Random Forest Classifier, they state that random decision forests easily adapt to non-linearity found in the data and therefore tend to predict better than linear regression.
+More specifically, ensemble learning algorithms like random forests are well suited for medium to large datasets.
+
+\end{itemize}
+
 Each of these algorithms used and their results are listed in the Results and Analysis Section \ref{sec:results-and-analysis} below.
@@
 
-In order for such to be processed in real-time, the Jupyter Notebook will start a PySpark session that is accessed via the local server running that accepts RESTful API calls with details of the server in Appendix \ref{appendix:code-api-server}.
+In order for each of the three ML models to be processed at the same time we crated a Jupyter Notebook that starts an Apache PySpark\footnote{PySpark: https://spark.apache.org/docs/latest/api/python/index.html} session that is accessed via the local server running that accepts RESTful API calls with details of the server in Appendix \ref{appendix:code-api-server}.
 
 \subsection{Evaluation Metrics and methodologies for measuring the performance of the ML models}
+\label{sec:evaluation-metrics-and-methodologies}
 The evaluation of ML models trained on the English, German and their respective extension datasets is a critical step in assessing their performance and reliability.
-We chose to use the MulticlassClassificationEvaluator\footnote{URL: https://spark.apache.org/docs/latest/api/python/reference/api/pyspark.ml.evaluation.MulticlassClassificationEvaluator.html} because is a well known tool in the field of natural language processing (NLP) research, specifically in multi class classification tasks such as Emotion Detection (ED).
-Generally this evaluator is used to evaluate the performance of machine learning (ML) models, using transformer-based architectures like BERT, RoBERTa, or DistilBERT\cite{Sanh2019-DistilBERT-AD}, with DistilBERT\cite{Sanh2019-DistilBERT-AD} being the one use in this research.
+We chose to use the MulticlassClassificationEvaluator\footnote{URL: https://spark.apache.org/docs/latest/api/python/reference/api/pyspark.ml.evaluation.MulticlassClassificationEvaluator.html} because is a well known tool in the field of natural language processing (NLP) research, specifically in multi class classification tasks such as emotion detection that has multiple labels associated with it; in our case eight (8) see Table \ref{table:dataset_labels} on Page \pageref{table:dataset_labels}.
+Generally this evaluator is used to evaluate the performance of machine learning (ML) models, using transformer-based architectures like BERT, RoBERTa, or DistilBERT\cite{Sanh2019-DistilBERT-AD}, with DistilBERT\cite{Sanh2019-DistilBERT-AD} being the one we used in this research project.
 
 The MulticlassClassificationEvaluator provides the essential evaluation metrics, including accuracy, precision, recall, and F1 score, for ED classifications.
-These metrics provide valuable insights into a model's ability to correctly classify emotions across multiple classes and language identification.
-Our research project is no different and the following that the assumptions are used to determine the effectiveness of the methodology of extending our datasets:
+These metrics provide valuable insights into a model's ability to correctly classify emotions across multiple classes and languages.
+Our research project is no different and the following measurements were used for our analysis:
 \begin{itemize}
@@
 It is calculated as the ratio of correctly classified instances to the total number of instances in the dataset.
-However, accuracy may not provide a complete picture, especially if the dataset is imbalanced.
 
-\item \textbf{Precision, Recall, and F1 Score:} Precision represents the proportion of correctly predicted emotions among all instances predicted as a specific emotion.
+\item \textbf{F1 Score:} Precision represents the proportion of correctly predicted emotions among all instances predicted as a specific emotion with respect to the labels.
 Recall measures the proportion of correctly predicted emotions among all instances that truly belong to a specific emotion.
-F1 Score ($F_{1}=\frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}$) combines precision and recall into a single metric, providing a balanced evaluation.
+With these two the F1 Score ($F_{1}=\frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}$) combines precision and recall into a single metric, providing a balanced evaluation.
 
-\item \textbf{Error Analysis:} Performing error analysis on the model's predictions helps identify specific patterns or instances where the model might struggle or make frequent mistakes.
-This analysis can guide improvements and provide insights for future model development.
 \end{itemize}
 
-When evaluating such models, several factors come into play.
-First and foremost, the choice of appropriate evaluation metrics is crucial, depending on the specific task at hand, such as text classification, sentiment analysis, or named entity recognition.
-Metrics like accuracy, precision, recall, F1 score, or area under the receiver operating characteristic curve (AUC-ROC) are commonly used to measure the models' performance.
+While evaluating the models we focused on the Accuracy and the F1 Score to illustrate the prediction results of the original versus extended datasets.
 
@@
 The following section lists the prediction rates of training and testing of the original and extended datasets both in English and German.
-Overall results seem to indicate that there was no and even negative results based on adding the extended data for Emotion Detection.
-In the below sub sections details on the results are listed in prose, tables, and formauls.
+Overall results seem to indicate that there was no measurable benefit and even negative results based on adding the extended data for Emotion Detection.
+In the below sub sections details on the results are listed in prose, tables, and formulas.
 
 \subsection{Prediction Results of Original \& Extended Datasets}
+Overall the prediction results from the trained models is disappointing.
+Using the Random Forrest Classifier described in Section \ref{sec:evaluation-metrics-and-methodologies} provided the best prediction results based.
+Oddly, the prediction results with the extended data where lower than using the original (non-extended) data for training.
 
-\ldots
+\subsubsection{English Dataset Results}
+The Table \ref{table:en_testing_results} lists the results of all three ML models, their accuracy and F1 Score based on the English original and extended datasets.
+The improvements from the original and extended datasets are listed in the last column.
+In all cases the improvement to using the extended datasets was negative.
 
-\subsection{Comparison of Prediction Rates of Original and Extended Language Datasets}
-\ldots
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | L | M | M | M | M | M | }
+    \hline
+    \multicolumn{6}{|c|}{Results of English Dataset} \\
+    \hline
+    &
+    \multicolumn{2}{c|}{Original Dataset} &
+    \multicolumn{3}{c|}{Extended Dataset} \\
+    & Accuracy & F1 Score & Accuracy & F1 Score & Improvement \\
+    \hline
+    Random Forrest & 
+    29.72\% &
+    20.32\% & 
+    28.77\%  &
+    19.47\% &
+    -3.20\% \\
+    \hline
+    Na\"ive Bayes & 
+    5.66\% &
+    6.92\% & 
+    3.77\%  &
+    4.11\% &
+    -33.39\% \\
+    \hline
+    GLM & 
+    25.94\% &
+    25.92\% & 
+    22.64\%  &
+    23.25\% &
+    -10.37\% \\
+    \hline
+\end{tabular}
+\caption{Model Testing Results on English Dataset}
+\label{table:en_testing_results}
+\end{table}
 
-\subsection{Analysis of the impact of Extending Datasets by Translation and Performance of the Combination of ML models}
-\ldots
+\subsubsection{German Dataset Results}
+The Table \ref{table:de_testing_results} lists the results of all three ML models, their accuracy and F1 Score based on the German original and translated from English to German extended datasets.
+The improvements from the original and extended datasets are listed in the last column.
+Interestingly the Randon Forrest Classifier was the only negative improvement whereas the GLM and Naive Bayes models had improvements.
+However it should be noted that the prediction rates are so low in the GLM and Naive Bayes that they should not really be considered improvements.
 
-\subsection{Discussion of the findings and their implications for improving Emotion Detection}
-\ldots
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | L | M | M | M | M | M | }
+    \hline
+    \multicolumn{6}{|c|}{Results of German Dataset} \\
+    \hline
+    &
+    \multicolumn{2}{c|}{Original Dataset} &
+    \multicolumn{3}{c|}{Extended Dataset} \\
+    & Accuracy & F1 Score & Accuracy & F1 Score & Improvement \\
+    \hline
+    Random Forrest & 
+    34.29\% &
+    17.51\% & 
+    18.57\%  &
+    13.44\% &
+    -45.84\% \\
+    \hline
+    Na\"ive Bayes & 
+    4.29\% &
+    4.31\% & 
+    6.67\%  &
+    4.21\% &
+    +55.48\% \\
+    \hline
+    GLM & 
+    11.90\% &
+    12.57\% & 
+    14.76\%  &
+    15.88\% &
+    +20.03\% \\
+    \hline
+\end{tabular}
+\caption{Model Testing Results on German Dataset}
+\label{table:de_testing_results}
+\end{table}
+
+The improvements on the accuracy are based on Equation \eqref{eq:improvement} listed below.
+It is a simple and straightforward way of comparing the Accuracy from the original trained dataset to the extended trained dataset for purposes of comparing the potential benefits of the translation.
+
+\begin{equation}
+Improvement(\%) = \dfrac{(Accuracy_{Ext} (\%) - Accuracy_{Org} (\%))}{Accuracy_{Org}(\%)}
+\label{eq:improvement}
+\end{equation}
+
+Further
+
+\subsection{Analysis of the impact of Extending Datasets by Translation}
+While extending datasets for machine learning training can often improve\cite{Sarker2021} model performance, it's important to recognize that this approach has its limitations, especially in ML models that can reach a limit on training data.
+In our case, relying on translation services for dataset expansion might not consistently yield better results when extending the data from translated text.
+These services, in our case Google Translate\footnote{Google Cloud Translate URL: https://cloud.google.com/translate} can introduce errors or inaccuracies in translations especially in the context of emotions of a particular text, leading to noisy or incorrect training data.
+Consequently, the models may learn from erroneous information, which did negatively impact our models performance on the ED tasks.
+
+\subsection{Findings and Implications for Improving Emotion Detection}
 
 \subsection{Conclusion and Future Work}
-\ldots
+It's crucial to ensure the quality and accuracy of extended data, especially when language translation is involved, to prevent introducing more noise than meaningful information into the training process.
+Although the research did not garner any significant improvements by translating the German to English as a means on extending the English dataset, it actually made it the prediction results worse.
+
+Future work includes testing and translating different models by using different translation services.
+Using Google Cloud's translation service was adequate for this research, however there are certainly better translation services that are available but are based on paid subscriptions.
 

### 2023-08-14 paper/appendix/code.tex
Renaming for Submission

@@
 \section{Appendix -- Application \& Code Details}
-Hello World - Application \& Code Details
-\subsection{Application}
-\label{appendix:code_application}
-Hello World - Application
+In this appendix we aim to supply details to the code and application blocks of this project.
+It should be noted that some of the code has been edited to fit on these pages and might not represent verbatim the code in the GIT repository.
+
+All code was developed and test using a MacBook PRO running \texttt{Ventura 13.4.1}.
+Since PySpark was developed for cluster processing it may seem counter intuitive to build PySpark code for a Mac; however the intent is that some of this research can be further used on clusters in the future.
+
+\subsection{PySpark Session Setup}
+\label{appendix:py-spark-setup}
+\begin{minted}[fontsize=\small]{python}
+# Setup Spark Session
+from pyspark.sql import SparkSession
+
+# Get Imports Needed
+from pyspark.sql.functions import col, udf, regexp_replace, lower
+from pyspark.sql import functions as F
+
+# Get Datatypes needed for DataFrame manipulation
+from pyspark.sql.types import IntegerType, StringType, ArrayType
+
+# Other Imports
+import re  # Import the "re" module for regular expressions
+
+# Setup Spark Session
+sc = SparkSession \
+        .builder \
+        .master("local[*]") \
+        .appName("Qualyfing-Exam") \
+        .getOrCreate()
+
+# Print Spark Version being run
+print("Spark V: ", sc.version)
+\end{minted}
 \clearpage
-
-\subsection{Code Details}
-Hello World - Code Details
 
@@
 
+Flask is a lightweight and simple to use web framework in Python, allowing developers to create web applications and APIs.
+Due to it's simplicity and components design, Flask allowed us to rapidly build and interactive web service fro demonstration purposes on this project.
+\begin{minted}[fontsize=\small]{python}
+# Setup API 
+from flask import Flask, request, jsonify
+from googletrans import Translator
+
+# Setup
+from transformers import DistilBertTokenizer
+from pyspark.ml.pipeline import PipelineModel
+
+# Initialize the AutoTokenizer
+tokenizer = DistilBertTokenizer.from_pretrained("distilbert-base-multilingual-cased")
+
+# Load Saved Models
+model_en_rfc = PipelineModel.load("./models/model_en_rfc.model")
+model_en_nb = PipelineModel.load("./models/model_en_nb.model")
+model_en_lr = PipelineModel.load("./models/model_en_lr.model")
+model_de_rfc = PipelineModel.load("./models/model_de_rfc.model")
+model_de_nb = PipelineModel.load("./models/model_de_nb.model")
+model_de_lr = PipelineModel.load("./models/model_de_lr.model")
+
+app = Flask(__name__)
+translator = Translator()
+
+# Setup Emotion Key Reverse
+emotions = {v: k for k, v in emotion_key.items()}
+
+@app.route('/predict', methods=['GET'])
+def transform_data():
+    sentence_en = request.args.get('sentence_en')
+
+    if not sentence_en:
+        return jsonify({'error': 'No Sentence (EN) provided!'})
+
+    sentence_de = translator.translate(sentence_en, dest="de").text
+    print(f"English: {sentence_en}")
+    print(f"German: {sentence_de}")
+
+    words_en = tokenizer.tokenize(sentence_en)
+    words_de = tokenizer.tokenize(sentence_de)
+
+    # Create Englsih & German Dataframe
+    df_en_api = sc.createDataFrame([(sentence_en, words_en)], ["sentence", "words"])
+    df_de_api = sc.createDataFrame([(sentence_de, words_de)], ["sentence", "words"])
+
+    predictions_en_rfc = model_en_rfc.transform(df_en_api)
+    predictions_en_nb = model_en_nb.transform(df_en_api)
+    predictions_en_lr = model_en_lr.transform(df_en_api)
+    predictions_de_rfc = model_de_rfc.transform(df_de_api)
+    predictions_de_nb = model_de_nb.transform(df_de_api)
+    predictions_de_lr = model_de_lr.transform(df_de_api)
+
+    #Build Response
+    resp = {
+        "sentence": {
+            "english":sentence_en,
+            "german":sentence_de,
+        },
+        "predictions": {
+            "en_rfc": emotions.get(predictions_en_rfc.collect()[0]["prediction"]),
+            "en_nb":  emotions.get(predictions_en_nb.collect()[0]["prediction"]),
+            "en_lr":  emotions.get(predictions_en_lr.collect()[0]["prediction"]),
+            "de_rfc": emotions.get(predictions_de_rfc.collect()[0]["prediction"]),
+            "de_nb":  emotions.get(predictions_de_nb.collect()[0]["prediction"]),
+            "de_lr":  emotions.get(predictions_de_lr.collect()[0]["prediction"]),
+        },
+    }
+
+    return jsonify(resp)
+
+if __name__ == '__main__':
+    app.run(host='127.0.0.1', port=8080)
+\end{minted}
 \clearpage

### 2023-08-14 paper/appendix/dataset.tex
Renaming for Submission

@@
 \begin{minted}[fontsize=\small]{python}
-# *********************
-# *** Setup Section ***
-# *********************
-
 import pandas as pd
 
-# Emotion DataFrame
 # This was predtermined by review of the emotion from English to German
@@
 
-\subsection{Details on English Dataset}
+\subsection{Details on English Dataset Parsing and Formatting}
 \label{appendix:dataset_english}
@@
 # *********************************
-
 import pandas as pd
@@
 df_en = pd.merge(df_en, df_emotions, on='emotion_en', how='left')
-
-# Add German Sentence Column
-df_en["sentence_de"] = ""
-
-# Randomly select 1500 rows
-df_en = df_en.sample(n=1500, random_state=2023)
+df_en["sentence_de"] = "" # Add German Sentence Column
+df_en = df_en.sample(n=1500, random_state=2023) # Randomly select 1500 rows
 
@@
 
-\subsection{Details on German Dataset}
+\subsection{Details on German Dataset Parsing and Formatting}
 \label{appendix:dataset_german}
@@
 
-# Merge Englsih Emotions onto German
+# Merge English Emotions onto German
 df_de = pd.merge(df_de, df_emotions, on='emotion_de', how='left')
-
-# Add English Sentence Column
-df_de["sentence_en"] = ""
-
-# Randomly select 1500 rows
-df_de = df_de.sample(n=1500, random_state=2023)
+df_de["sentence_en"] = "" # Add English Sentence Column
+df_de = df_de.sample(n=1500, random_state=2023) # Randomly select 1500 rows
 

### 2023-08-14 paper/main.tex
Renaming for Submission

@@
 \newcolumntype{M}{>{\centering\arraybackslash}m{0.85in}}
+\usepackage{tablefootnote}
 
@@
 \begin{abstract}
-This Qualifying Exam\footnote{PhD students of MTSU Computational and Data Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages specifically  English and German.
-The research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis research and projects.
+This Qualifying Exam\footnote{PhD students of MTSU Computational and Data Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages.
+Our research focused on  English and German datasets.
+This research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis research and projects, essence the lack of data to better train ML models is trying to be addressed.
 
-By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) the hope is that better prediction rates can be achieved in ED analysis applications.
-Furthermore, translating English to German to extend the German dataset and training another ML model in parallel using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates by dual processing a sentence at the same time.
+By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) become larger and the hope is that better prediction rates can be achieved in ED analysis applications due to ML training having more data for learning.
+Furthermore, translating English to German to extend the German dataset and training another ML model using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates by dual processing a sentence at the same time.
 
 In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, translated, and two ML models built that can be accessed via an API to get a prediction on an input (POST or GET) sentence both in German or English.
+
+Sadly the research and results point to the realization that extending the datasets by translation did not improve prediction rates of English or German.
+Some minor gains can be achieve by accessing both English and German models at the same time via the RESTful API but unfortunately the benefits may not really support the efforts.
 \end{abstract}
@@
 Emotion detection has gained significant interest in recent years as a means to understand human behavior and improve communication effectiveness.
-By analyzing text or speech, these detection processes can accurately identify and classify emotions expressed by individuals or groups, often times in real-time.
+By analyzing text, speech or facial expressions, these detection processes try to accurately identify and classify emotions expressed by individuals or groups, often times in real-time.
 
-By employing a variety of ED processes, including natural language processing and machine learning, applications can detect emotions with high precision.
+By employing a variety of ED processes, including natural language processing (NLP) and machine learning (ML), applications with the right amount of data can detect emotions with high precision.
 These ED systems find applications in diverse fields, such as sentiment analysis, customer feedback analysis, voting and news stances, and mental health monitoring, which all aim at contributing to a deeper understanding of human emotions and their impact on various aspects of life.
@@
 One well-known model is the Plutchik's Wheel of Emotions Figure \ref{fig:plutchik-wheel}, which proposes eight (8) primary emotions and their contrasting pairs.
-In this paper we will be focusing on the basic eight emotions and not on their sub-emotions.
+In this research paper we will be focusing on the basic eight emotions and not on their sub-emotions or contrasting pairs.
+This allows the ML models being trained and built to be a multi-classification model with eight label options.
 
-Below are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp}, a detailed reference of emotions used in this research can be found in Table \ref{table:dataset_labels} on page~\pageref{table:dataset_labels}.
+Listed below are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp} that we will be focusing on both in English and German, a detailed reference of emotions used in this research can be found in Table \ref{table:dataset_labels} on page~\pageref{table:dataset_labels}.
 
@@
 
-It's important to note that emotions are complex and multifaceted, and different models may propose alternative categorizations or variations in the number of primary emotions.
+It's important to note that emotions are complex and multifaceted, and different emotion models or frameworks may propose alternative categorizations or variations in the number of primary emotions.
 However for the scope of this research project the most common eight (8) emotions are used as described above.
+This allows the ML models being trained and built to be a multi-classification model with eight label options.
 
-By aligning the emotional labels across the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets being used in this research all eight listed above were used.
-Details on the linkage of the two datasets is described in Section \ref{sec:data-procurement-process} with an emphasis on the Table \ref{table:dataset_labels}'s cross reference on page~\pageref{table:dataset_labels}.
+By aligning the emotional labels across the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets that are being used in this research all eight listed above were used.
+Details on the linkage of the two datasets and their counts are described in Section \ref{sec:data-procurement-process} with an emphasis on Table \ref{table:dataset_labels}'s cross reference on page~\pageref{table:dataset_labels} of German and English languages.
 
-\subsection{Significance of Emotion Detection (ED) in Machine Learning}
+\subsection{Significance of Emotion Detection in Machine Learning}
 Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions (chatbots\cite{chatbot-cognitive-awareness}) to just name a few.
@@
 By accurately identifying and understanding emotions from text data, ML applications can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.
-As an example, the primary goals of chatbots are entertainment, social contact, and novelty interaction, with a strong emphasis on productivity, scale-ability and cost reductions.
-In chatbots, there has been a immense requirement to simulate human-like characteristics and behavior during human-machine interactions\cite{emotion-detection-literature-review}.
+
+As an example, the primary goals of chat-bots are entertainment, social contact, and novelty interaction, with a strong emphasis on productivity, scale-ability and cost reductions.
+In chat-bots, there has been a immense requirement to simulate human-like characteristics and behavior during human-machine interactions\cite{emotion-detection-literature-review}.
 Per Kusal et al.
-\cite{chatbot-cognitive-awareness} these chatbots also improve customer engagement by offering friendliness, comfortness, flexibility, and efficient assistance.
-Chatbots should provide customers with more engaging responses, directly addressing their issues.
-The goal of deploying ED applications is for users perceive chatbots as companions rather than simple assistants.
+\cite{chatbot-cognitive-awareness} these chat-bots also improve customer engagement by offering friendliness, comforters, flexibility, and efficient assistance.
+Chat-bots should provide customers with more engaging responses, directly addressing their issues or questions.
+The goal of deploying ED applications is for users perceive chat-bots as companions rather than simple assistants, which in turn helps the overall process of getting to a positive result.
 Per Adikari et al.
-the majority of user requests are emotional rather than informative.
-Chatbots gained the capacity to respond emotionally to customers due to machine learning and sentiment analysis evolution\cite{chatbot-cognitive-awareness}.
+the majority of user requests are emotional rather than informative; therefore the need that chat-bots gain the capacity to respond emotionally to customers due to machine learning and sentiment analysis evolution\cite{chatbot-cognitive-awareness} is very important and yields for many applications that employee ED a way to improve interactions with machines.
 
@@
 Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions.
-This in turn contributes to the advancement of artificial intelligence, natural language processing and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions}.
+This in turn contributes to the advancement of artificial intelligence, natural language processing (NLP) and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions} to go about our daily lives.
 
-Lastly, in Chen et al.\cite{research-emotion-recognition-for-online-learning} paper based on emotion detection for online learning, it states that with emotion recognition research, we need to focus on accuracy and real-time performance in order to apply emotional recognition based on physiological signals to solve practical problems; they key part of their claim is that in order for ED to be practical real-time methods needs to be used.
+Lastly, in Chen et al.\cite{research-emotion-recognition-for-online-learning} paper based on emotion detection for online learning, it states interestingly that with emotion recognition research, we need to focus on accuracy and real-time performance in order to apply ED based on physiological signals to solve practical problems; the key part of their claim is that in order for ED to be practical, real-time methods needs to be used and deployed for maximum results.
 
-\subsection{ED challenges: Lack of Labeled Datasets and Language Fragmentation}
-ED is an immensely growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}.
-The primary reason of the lack of data stems from the emotion detection data requiring primarily supervised learning dataset, which require time and effort to procure, and mostly created and labeled by humans.
-To compound on the issue of the lack of data; the many datasets that are available are in multiple languages.
-A significant amount of time took place to find the German labeled
+\subsection{Challenges: Lack of Labeled Datasets and Language Fragmentation}
+Emotion Detection is an immensely growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}.
+The primary reason of the lack of data stems from the face that emotion detection data requires primarily supervised learning data.
+In in turn requires time and effort to construct, catalog, and procure,; this all is mostly created and labeled by humans.
+To compound on the issue of the lack of labeled data; the many datasets that are available are in many cases in multiple languages - not all are in English since emotions that are linked to text are contextual in nature.
+for this research a a significant amount of time took place to find the German labeled data, which eventually was found online by google searching.
 
@@
 This paper is significant because an entire industry of automated emotion reading (Text, Image, Video, and Sound) technologies is quickly emerging \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
-The market for emotion recognition software is forecast to reach at least \$3.8 billion by 2025 \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
-And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, customer service and a multiple of others, it is fitting to work on research to solve issues with prediction rates and lack of teaching / learning data.
+The market for emotion recognition and detection software is forecast to reach at least \$3.8 billion by 2025 \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
+And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, customer service, news and stance analysis, and a multiple of others, it is fitting to work on research to solve and / or improve issues in regards to prediction rates and lack of teaching / learning data.
 
 With the before mentioned issues regarding the quantity (lack of) and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED.
-The hope is that by use of translation that the predictability of ED can be improved by use of real-time language conversion and processing.
+The aim of the research is that by use of translation that the predictability of ED can be improved by extending an English dataset with translations from German and also process look-ups in real-time language conversion and processing.
 
@@
 \begin{itemize}
-\item Can by translating English data to German and extending an original German dataset increase the predictability of the ED model?
-\item Can by translating German data to English and extending an original English dataset increase the predictability of the ED model?
-\item Can by translating in real-time an input to multiple languages improve the predictability based on the combined output of two model's results.
+\item Can by translating English data to German and extending an original German dataset increase the predictability of German ED model?
+\item Can by translating German data to English and extending an original English dataset increase the predictability of English ED model?
+\item Can by translating in real-time an input to multiple languages improve the predictability based on the combined output of two models that were trained in English and German to better the prediction results.
 \end{itemize}
+
+With these three (3) objectives our approach will be to procure, translate, train, and evaluate benefits of extending datasets by translation to improve emotion detection.
 
@@
 \section{Literature Review}
-To ensure comprehensive coverage of the latest developments and research in emotion detection and analysis for this research paper, consideration and review was only given to incorporating research papers and online news articles that were published after the year 2015, thereby providing an up-to-date perspective on the subject matter.
-By implementing the 2015 boundary, our methodology aims to provide a more up-to-date outlook on the subject matter at hand and what areas the ED field in moving in in respect to ML applications and processes.
+To ensure comprehensive coverage of the latest developments and research in emotion detection and analysis for this research paper, consideration and review was only given to incorporating research papers and online news articles that were published after the year 2015, thereby providing an up-to-date perspective on the subject matter at hand.
+
+We believe by implementing the 2015 boundary, our methodology and results aim to provide a more up-to-date outlook on the subject matter at hand and what areas the emotion detection field is moving in with respect to machine learning research, applications and processes.
 
@@
 Emotion Detection (ED) holds significant importance in various domains, including psychology, social sciences, and human-computer interaction.
-It enables the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states.
-In Chowanda et.
-al.
-paper\cite{CHOWANDA-2021821} they believe that "Emotions hold a paramount role in the conversation, as it expresses context to the conversation."
+It enables the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states both in real-time but also as a type barometer for measuring sentiments from the past versus the current time.
+In Chowanda et al.
+paper\cite{CHOWANDA-2021821} they believe that "Emotions hold a paramount role in the conversation, as it expresses context to the conversation.", this means that emotions are a part of a conversation and with that are needed to ensure valid analysis of a conversation.
 
-Furthermore, emotions can differ across age groups, genders, cultures, and languages.
+Furthermore Kapoor et al.
+\cite{KAPOOR2023120882} states that, emotions can differ across age groups, genders, cultures, and languages.
 Including data from diverse populations helps in developing inclusive and culturally sensitive ED models.
@@
 \subsection{Lack of Labeled Datasets and Fragmentation caused by Different Languages}
+The lack of diverse and large-scale data hinders the development and training of accurate ED models, limiting their effectiveness and general ability across different contexts and populations.
+In recent research, efforts have been made to address these limitations by exploring new techniques for emotion detection.
 
-The lack of diverse and large-scale data hinders the development and training of accurate ED models, limiting their effectiveness and general ability across different contexts and populations.
-
-In recent research, efforts have been made to address these limitations by exploring new techniques for emotion detection.
 For instance, a study by Kapoor\cite{KAPOOR2023120882} et al.
 proposes analyzing the entire spectrum of the language source to predict emotional changes.
-This provided some help, but quickly established itself as time consuming and need of expertise in the subject matters.
+This provided some help, but quickly established itself as time consuming task and need of expertise in the subject matters that could not outweigh the results.
 Overcoming the scarcity of labeled datasets and embracing innovative approaches like the before mentioned study by Kapoor\cite{KAPOOR2023120882} et al.
 and their analysis can contribute to the advancement of ED research.
-Such advancements hold the potential to enhance the accuracy and applicability of emotion detection models, enabling a deeper understanding of human emotions and more effective responses in various domains.
+Such advancements hold the potential to enhance the accuracy and applicability of emotion detection models, enabling a deeper understanding of human emotions and more effective responses in various domains, although the cost of such might be prohibitive at this time.
 
-The paper by Mascarell\cite{mascarell-etal-2021-stance} et al.
+In addition, the paper by Mascarell\cite{mascarell-etal-2021-stance} et al.
 states that unlike images language operate very differently, therefore the lack of labeled datasets poses a significant challenge in text-based Emotion Detection (ED) analysis as it limits the availability of reliable training data for ML models.
@@
 
-Additionally, Kusal et al.
+And lastly, a paper by Kusal et al.
 further stipulates \cite{kusal} that the fragmentation caused by different languages further exacerbates the issue, as it reduces the size and diversity of data available for training, resulting in limited cross-lingual generalization and potentially biased models.
-Overcoming these challenges is crucial for advancing ED research and improving the accuracy and applicability of emotion detection models across various languages and cultures.
+
+Overcoming these challenges, lack of data and language fragmentation, is crucial for advancing emotion detection and analysis research for improving the accuracy, speed, and applicability of ML models across various languages and cultures.
 
@@
 \label{sec:data-procurement-process}
-The data procurement was relatively straight forward and in essence entailed using Google search terms for the Emotion Detection in English and German.
+The data procurement was relatively straight forward and in essence entailed using Google search terms for the Emotion Detection in English and German text languages.
 Some of the main German data was obtained from the dataset built by ETH's Emotion and Stance Detection for German Text \cite{mascarell-etal-2021-stance} which also hosts a good website with their findings \footnote{https://mtc.ethz.ch/research/natural-language-processing/emotion-stance.html}.
@@
 The English dataset contains over 38,000 tweets and their corresponding emotion labeled.
-Similarly the German contains over 12,000 sentences and their respective emotions associated.
+Unfortunately the German dataset only contains over 2,000 sentences and their respective emotions associated.
+this imbalance of the two dataset sizes will be corrected in the subsequent sections.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | c | c | c | }
+    \hline
+    \multicolumn{3}{|c|}{File Details} \\
+    \hline
+    Name & Row Count & Type \\
+    \hline
+    English & 38,000 & CSV  \\
+    German  &  2,500 & JSON \\
+    \hline
+\end{tabular}
+\caption{File details of English \& German data.}
+\label{table:file-details}
+\end{table}
 
 An important part of the process is to make sure we match the English emotions to those of the German language emotions.
-Since the tow languages are dissimilar we opted to match exact emotions to each other that can be seen in Table \ref{table:dataset_labels} below.
+Since the two languages are dissimilar we opted to match exact emotions to each other that can be seen in Table \ref{table:dataset_labels} below.
 
@@
     \hline
-    \multicolumn{2}{|c|}{English} & \multicolumn{2}{|c|}{German} & \multirow{2}{*}{Used} \\
+    \multicolumn{2}{|c|}{English} & \multicolumn{2}{c|}{German} & \multirow{2}{*}{Used} \\
     Name & Count & Name & Count \\
@@
 
-The procurement of two ED datasets is needed.
+The processing of two ED datasets to be synchronized to the emotion is important.
 The first will be in English and the second in German.
 Each dataset will be translated into the other’s language for extending purposes.
+And each row will hold the English and German sentence.
 As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.
+By this simple process the English dataset got larger for training purposes.
 
@@
     \centering
-    \includegraphics[width=0.35\textwidth]{Dataset-Extension}
+    \includegraphics[width=0.4\textwidth]{Dataset-Extension}
     \caption{Example of English dataset extension by translating from German.}
@@
 
-The reason for selecting German was based on the author\footnote{Richard Hoehn} being fluent in English and German and hence being able to easily traverse the two languages while working on the code for dataset translations and formatting.
+The reason for selecting German in this research project was based on the author\footnote{Richard Hoehn} being fluent in English and German and hence being able to easily traverse the two languages while working on the code for dataset translations and formatting.
+It was much easier to create the emotion cross reference Table \ref{table:dataset_labels} and also check the translation results due to the dual language benefit.
 
@@
 \subsection{Development of a translation application for converting different languages into English and vice versa}
-Translation of the emotion datasets was performed with Google Translator based on the Medium article\cite{Nidhaloff_how_to_translate_text_with_python} in which the author describes option for simple API translations.
-As an example the following is a brief introduction to Google Translator's API:
+Translation of the emotion datasets was performed with Google Translator based on the Medium article\cite{Nidhaloff_how_to_translate_text_with_python} in which the author describes options for simple API translations.
+Google Cloud Translation was use mainly due to it's free nature and ease of use for this project.
+There are however many other text based API translation services that could have been used and might yield in some circumstance even better results.
+
+As an example the following is a brief introduction to Google Translator's API with the code snippet below depicting the ease of use to translate text via an API.
 
 \begin{quote}
-\begin{minted}[fontsize=\small]{python}
+\begin{minted}[samepage, fontsize=\small]{python}
+#
 # Using Deep Translator to leveage Google Translate
@@
 print(translated_back) # Chocolate milk tastes much better through a straw.
+
 \end{minted}
@@
 
-Based on the above example the CSV and JSON datasets were individually loaded, parsed, and translated to their respective counter-part languages.
-Details of the application can be found in the Appendix \ref{appendix:dataset_english} for English and Appendix \ref{appendix:dataset_german} for the German.
-We decided to only use 1,500 row for each language so since the German (DE\footnote{DE = German}) dataset was not much larger than that.
-This meant that the English (EN\footnote{EN = Enlgish}) datasets was randomly paired down to 1,500 to match that of the German.
+Based on the above translation example the CSV (English) and JSON (German) datasets were individually loaded, parsed, and translated to their respective counter-part languages.
+Details of the application can be found in the Appendix \ref{appendix:dataset_english} for English and Appendix \ref{appendix:dataset_german} for the German processing.
+We decided to only use 1,500 rows from each language.
+This decision was made since the German (DE\footnote{DE = German}) dataset was not much larger than that once filtering and processing was completed.
+This meant that the English (EN\footnote{EN = English}) dataset was randomly paired down to 1,500 to match that of the German.
 
@@
 \subsection{Training Multiple ML models with PySpark}
+After the completion of translating and processing the Englsih and German datasets we come to training our models.
+Using the datasets from Tables \ref{table:pd_de_translated}, \ref{table:pd_en_translated}, and \ref{table:ml_models} that list an approximate $\sim$1,275 and $\sim$2,775 features for training.
 
- Using the datasets from Tables \ref{table:pd_de_translated}, \ref{table:pd_en_translated}, and \ref{table:ml_models} that list an approximate $\sim$1,275 and $\sim$2,775 features for training.
+It is important to note that in order to accurately test and compare the original and the extended datasets to each other we made sure that the test data was the same.
+to do this we randomly extracted 15\% of the original datasets (English and German) before extending the original with the translated datasets.
+
+\begin{quote}
+\begin{minted}[samepage, fontsize=\small]{python}
+#
+# Split the English and German Dataframes for Training and Testing
+# We are using a 15/85 Split
+df_en_train, df_en_test = df_en.randomSplit([0.85, 0.15], seed=2023)
+df_de_train, df_de_test = df_de.randomSplit([0.85, 0.15], seed=2023)
+
+# Create the Extended Dataframe with Translated Data
+# It is important to add the translated to the training data and not all of it
+# Using *df_en_train.columns ensure that the columns are matched with the union
+df_en_train_extended = df_en_train.union(df_de.select(*df_en_train.columns))
+df_de_train_extended = df_de_train.union(df_en.select(*df_de_train.columns))
+
+\end{minted}
+\end{quote}
+
+In Chowanda et al.
+paper\cite{CHOWANDA-2021821} on Text-based Emotion Techniques multiple algorithms were proposed, our research focused on just three (3) out of the six from the paper since the others had lower prediction rates based on the paper\'s conclusions.
 The algorithms tested in this research project ranged from the traditional machine learning to deep learning techniques, they are:
@@
 Each of these algorithms used and their results are listed in the Results and Analysis Section \ref{sec:results-and-analysis} below.
-In Chowanda et al.
-paper\cite{CHOWANDA-2021821} on Text-based Emotion Techniques multiple algorithms were proposed, this research focused on just three out of their six since the others had lower prediction rates based on Chowanda et al.
-paper\cite{CHOWANDA-2021821} paper.
 
@@
     225 same as Model "C" \\
-
     \hline
@@
 
+The results of the testing resulted in some interesting parts that can be viewed in Table \ref{table:ml-prediction-result-versus-test-data}.
+We grouped the label by count from the test data (both English and German) and compared the predicted result counts to those.
+It became very clear that there was a strong bias on the GLM (Linear Regression) model that essentially proposed that all sentences evalued to the same output both in English and German.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | M | M | M | M | M | M | }
+    \hline
+    \multicolumn{6}{|c|}{ML Models Test versus Test Result Counts} \\
+    \multicolumn{6}{|c|}{} \\
+    \multicolumn{6}{|c|}{English} \\
+    \hline
+    ID & 
+    Label &
+    Test Data &
+    NB Res & 
+    GLM Res  &
+    RFC Res \\
+    \hline
+    0 & boredom     &  0 & 25 \tablefootnote{In the NB a result that cannot be matched is a zero label} &   0 &   0 \\
+    1 & love        & 26 &  8 &   3 &  11 \\
+    2 & Relief      &  0 & 33 &   0 &   0 \\
+    3 & Fun         &  0 &  5 &   0 &   0 \\
+    4 & hate        &  9 & 31 &   0 &   0 \\
+    5 & neutral     & 50 & 21 & 152 & 101 \\ 
+    6 & anger       &  0 & 36 &   0 &   0 \\
+    7 & happiness   & 39 & 44 &   5 &  17 \\
+    8 & surprise    & 10 &  9 &   0 &   1 \\
+    9 & sadness     & 29 &  0 &   2 &  19 \\
+    10 & worry      & 46 &  0 &  50 &  63 \\
+    11 & enthusiasm &  3 &  0 &   0 &   0 \\
+    \hline
+    \multicolumn{6}{|c|}{} \\
+    \multicolumn{6}{|c|}{German} \\
+    \hline
+    ID & 
+    Label &
+    Test Data &
+    NB Res & 
+    GLM Res  &
+    RFC Res \\
+    \hline
+    0 & boredom     &  0 & 34 &   0 &   0 \\
+    1 & love        & 25 &  5 &   0 &  13 \\
+    2 & Relief      &  0 & 28 &   0 &   0 \\
+    3 & Fun         &  0 & 16 &   0 &   0 \\
+    4 & hate        &  5 & 16 &   0 &   2 \\
+    5 & neutral     & 30 & 24 &   0 &   5 \\ 
+    6 & anger       & 23 & 20 &   1 &   7 \\
+    7 & happiness   & 10 & 20 &   0 &   5 \\
+    8 & surprise    & 24 & 47 &   0 &  11 \\
+    9 & sadness     & 12 &  0 &   0 &   2 \\
+    10 & worry      &  9 &  0 & 209 &   4 \\
+    11 & enthusiasm & 72 &  0 &   0 & 161 \\
+    \hline
+\end{tabular}
+\caption{Prediction Results Count compared to Test Data Label Counts}
+\label{table:ml-prediction-result-versus-test-data}
+\end{table}
+
 In order for each of the three ML models to be processed at the same time we crated a Jupyter Notebook that starts an Apache PySpark\footnote{PySpark: https://spark.apache.org/docs/latest/api/python/index.html} session that is accessed via the local server running that accepts RESTful API calls with details of the server in Appendix \ref{appendix:code-api-server}.
 
-\subsection{Evaluation Metrics and methodologies for measuring the performance of the ML models}
+\subsection{Evaluation Metrics and Methodologies for Measuring the Performance}
 \label{sec:evaluation-metrics-and-methodologies}
 The evaluation of ML models trained on the English, German and their respective extension datasets is a critical step in assessing their performance and reliability.
-We chose to use the MulticlassClassificationEvaluator\footnote{URL: https://spark.apache.org/docs/latest/api/python/reference/api/pyspark.ml.evaluation.MulticlassClassificationEvaluator.html} because is a well known tool in the field of natural language processing (NLP) research, specifically in multi class classification tasks such as emotion detection that has multiple labels associated with it; in our case eight (8) see Table \ref{table:dataset_labels} on Page \pageref{table:dataset_labels}.
+We chose to use the MulticlassClassificationEvaluator\footnote{URL: https://spark.apache.org/docs/latest/api/python/reference/api/pyspark.ml.evaluation.MulticlassClassificationEvaluator.html} because is a well known tool in the field of natural language processing (NLP) research, specifically in multi-class classification tasks such as emotion detection that has multiple labels associated with it; in our case eight (8) see Table \ref{table:dataset_labels} on Page \pageref{table:dataset_labels}.
 Generally this evaluator is used to evaluate the performance of machine learning (ML) models, using transformer-based architectures like BERT, RoBERTa, or DistilBERT\cite{Sanh2019-DistilBERT-AD}, with DistilBERT\cite{Sanh2019-DistilBERT-AD} being the one we used in this research project.
 
-The MulticlassClassificationEvaluator provides the essential evaluation metrics, including accuracy, precision, recall, and F1 score, for ED classifications.
+The MulticlassClassificationEvaluator provides the essential evaluation metrics, including accuracy, Precision (eq: \ref{eq:precision}), Recall (eq: \ref{eq:recall}), and F1 Score (eq: \ref{eq:f1-score}) for ED classifications.
 These metrics provide valuable insights into a model's ability to correctly classify emotions across multiple classes and languages.
@@
 Recall measures the proportion of correctly predicted emotions among all instances that truly belong to a specific emotion.
-With these two the F1 Score ($F_{1}=\frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}$) combines precision and recall into a single metric, providing a balanced evaluation.
+With these two the F1 Score ($F1_{Score}=\frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}$) combines precision and recall into a single metric, providing a balanced evaluation.
 
@@
 
-While evaluating the models we focused on the Accuracy and the F1 Score to illustrate the prediction results of the original versus extended datasets.
+While evaluating the models we focused on the Accuracy and the F1 Score to illustrate the prediction results of the original versus extended datasets, and also compared the different algorithms to each other.
+As an added example simply using the accuracy to evaluate a model can be miss-leading and therefore in our case using the F1 Score, that is directly correlated to the Precision (eq: \ref{eq:precision} and Recall (eq: \ref{eq:recall}) is immensely important to understanding the ones models and finding the best of for the task of Emotion Detection using Original and Extended datasets.
 
 \subsection{Written \& Oral Presentation of Research Finding}
-In conclusion of the project a research paper and oral presentation is required to pass MTSU Computation Science PhD Qualifying Exam.
-The listing the datasets used, code for the application, and details of the research project process in this research paper is openly available and hosted on a GitHub repository, allowing for easy access, reproducible, and collaborative engagement with the research findings.
-The repository is open and available under: \texttt{https://github.com/richardhoehn/mtsu.coms.qualifying-exam}\cite{Hoehn_Improving_Emotion_Detection_2023}
+In conclusion of the project a research paper and oral presentation is required to pass the  MTSU Computation Data Science PhD Qualifying Exam.
+The listing of all the datasets used, code for the application, and details of the research project process in this research paper is openly available and hosted on a GitHub repository, allowing for easy access, reproducible, and collaborative engagement with the research findings.
+The repository is open and available under: \texttt{https://github.com/richardhoehn/mtsu.coms.qualifying-exam}\cite{Hoehn_Improving_Emotion_Detection_2023}.
+In addition of all the code being listed in a public repository the paper was comprised and build using \LaTeX\footnote{LaTeX UR: https://www.latex-project.org/}.
+
+Using \LaTeX\;for this paper provides numerous benefits, some of which are portability, source control (due to plain-text nature), bibliography management, and last but not least the ability to write clean mathematical equations.
 
@@
 The following section lists the prediction rates of training and testing of the original and extended datasets both in English and German.
-Overall results seem to indicate that there was no measurable benefit and even negative results based on adding the extended data for Emotion Detection.
-In the below sub sections details on the results are listed in prose, tables, and formulas.
+Overall results seem to indicate that there was no measurable benefit and even negative outcomes based on adding the extended data for Emotion Detection.
+In the below sub sections details on the results are listed by the common standards of using the below equations \ref{eq:accuracy}, \ref{eq:precision}, \ref{eq:recall}, and \ref{eq:f1-score} for Accuracy, Precision, Recall, and the F1 Score.
+It should be noted that ED is a Multi-class classification problem, which means that the machine learning process assigns the predictions from the sentence data to one of several predefined emotion categories (see Table \ref{table:dataset_labels}).
 
@@
 Overall the prediction results from the trained models is disappointing.
-Using the Random Forrest Classifier described in Section \ref{sec:evaluation-metrics-and-methodologies} provided the best prediction results based.
+Using the Random Forrest Classifier described in Section \ref{sec:evaluation-metrics-and-methodologies} provided the best prediction results based on a combination of the Accuracy and the F1 Score.
+It should be noted although the accuracy may be slightly better on the GLM models for both English and German, if one reviews the F1 Score, the RFC model performs better.
+
 Oddly, the prediction results with the extended data where lower than using the original (non-extended) data for training.
+In the analysis and conclusions section \ref{sec:analysis-of-impact-of-extending-datasets} below details on why this may be the case are discussed in detail.
+
+\subsubsection{Formulas and Equations used in Analysis}
+The accuracy (eq: \ref{eq:accuracy}) of any machine learning model is a clear and easily comprehend-able way to understand the performance of a model.
+However, it is important to consider the context and potential trade-offs with other metrics, as high accuracy alone might not capture the full picture of a model's performance.
+\begin{equation}
+Accuracy = \frac{TP+TN}{TP+TN+FP+FN}
+\label{eq:accuracy}
+\end{equation}
+
+The Precision (eq: \ref{eq:precision}) focuses in detail on the accuracy of positive predictions (TP)\footnote{Positive Predictions = TP (True Positive)} made by the model.
+Per Google Cloud's definition the precision answers: What proportion of positive predictions where actually correct?
+\begin{equation}
+Precision = \frac{TP}{TP+FP}
+\label{eq:precision}
+\end{equation}
+Since our project is a multi-class classification project we will need to actually use Micro Averaging for the Precision.
+With this in mind we extend our simple equation \ref{eq:precision} of precision to incorporate micro averaging is illustrated in equation \ref{eq:precision_micro_avg}.
+\begin{equation}
+Precision_{MicroAvg} = \frac{(TP_1 + TP_2 + \ldots + TP_n)}{(TP_1 + TP_2 + \ldots + TP_n + FP_1 + FP_2 + \ldots + FP_n)}
+\label{eq:precision_micro_avg}
+\end{equation}
+
+The Recall (eq: \ref{eq:recall}) focuses on the in detail on the accuracy of positive predictions\footnote{Negative Predictions = FN (False Negative)} made by the model.
+Per Google Cloud's definition: The recall attempts to answer: What proportion of actual positives.
+
+\begin{equation}
+Recall = \frac{TP}{TP+FN}
+\label{eq:recall}
+\end{equation}
+
+Since our project is a multi-class classification project we will need to actually use Micro Averaging for the Recall in a similar fashion to how we used it on Precision.
+With this in mind we extend our simple equation \ref{eq:recall} of recall to incorporate micro averaging is illustrated in equation \ref{eq:recall_micro_avg}.
+
+\begin{equation}
+Recall_{MicroAvg} = \frac{(TP_1 + TP_2 + \ldots + TP_n)}{(TP_1 + TP_2 + \ldots + TP_n + FP_1 + FP_2 + \ldots + FP_n)}
+\label{eq:recall_micro_avg}
+\end{equation}
+
+The F1 score (eq: \ref{eq:f1-score}) in our project is important to assess the balance between precision and recall.
+In addition, the F1 score is especially relevant in multi-class classification scenarios where there are more than two classes.
+
+\begin{equation}
+F1_{Score} = \frac{2*Precision*Recall}{Precision+Recall}
+\label{eq:f1-score}
+\end{equation}
+
+Lastly, we used micro averaging for the Recall, Precision, and F1 Score due to fact it give each label (Emotion see Table \ref{table:dataset_labels}) equal weight regardless of the class label and the number of labels counted.
 
 \subsubsection{English Dataset Results}
+\label{sec:english-dataset-results}
 The Table \ref{table:en_testing_results} lists the results of all three ML models, their accuracy and F1 Score based on the English original and extended datasets.
@@
     29.72\% &
-    20.32\% & 
-    28.77\%  &
-    19.47\% &
-    -3.20\% \\
+    25.83\% & 
+    21.23\%  &
+    19.43\% &
+    -28.56\% \\
     \hline
@@
     3.77\%  &
-    4.11\% &
+    4.14\% &
     -33.39\% \\
     \hline
-    GLM & 
-    25.94\% &
-    25.92\% & 
-    22.64\%  &
-    23.25\% &
-    -10.37\% \\
+    GLM (lr) & 
+    32.08\% &
+    23.68\% & 
+    25.47\%  &
+    16.19\% &
+    -19.95\% \\
     \hline
@@
 \subsubsection{German Dataset Results}
+\label{sec:german-dataset-results}
 The Table \ref{table:de_testing_results} lists the results of all three ML models, their accuracy and F1 Score based on the German original and translated from English to German extended datasets.
 The improvements from the original and extended datasets are listed in the last column.
-Interestingly the Randon Forrest Classifier was the only negative improvement whereas the GLM and Naive Bayes models had improvements.
+Interestingly the Random Forrest Classifier was the only negative improvement whereas the GLM and Naive Bayes models had improvements.
 However it should be noted that the prediction rates are so low in the GLM and Naive Bayes that they should not really be considered improvements.
@@
     Random Forrest & 
-    34.29\% &
-    17.51\% & 
-    18.57\%  &
-    13.44\% &
-    -45.84\% \\
+    28.10\% &
+    16.84\% & 
+    25.71\%  &
+    18.56\% &
+    -8.50\% \\
     \hline
@@
     \hline
-    GLM & 
-    11.90\% &
-    12.57\% & 
-    14.76\%  &
-    15.88\% &
-    +20.03\% \\
+    GLM (lr) & 
+    34.29\% &
+    17.57\% & 
+    24.76\%  &
+    18.27\% &
+    -27.79\% \\
     \hline
@@
 The improvements on the accuracy are based on Equation \eqref{eq:improvement} listed below.
-It is a simple and straightforward way of comparing the Accuracy from the original trained dataset to the extended trained dataset for purposes of comparing the potential benefits of the translation.
+It is a simple and straightforward way of comparing the Accuracy from the original trained dataset to the extended trained dataset for purposes of comparing the potential benefits of extending the training datasets by translation.
 
@@
 
-Further
+It should be noted that although the improvement values are fairly large negatives those come from the fact that the accuracy is small and in most all cases the accuracy is around the 25 - 30\% values and a change of just a few percent can have a significant impact on the improvement values.
 
 \subsection{Analysis of the impact of Extending Datasets by Translation}
+\label{sec:analysis-of-impact-of-extending-datasets}
 While extending datasets for machine learning training can often improve\cite{Sarker2021} model performance, it's important to recognize that this approach has its limitations, especially in ML models that can reach a limit on training data.
 In our case, relying on translation services for dataset expansion might not consistently yield better results when extending the data from translated text.
-These services, in our case Google Translate\footnote{Google Cloud Translate URL: https://cloud.google.com/translate} can introduce errors or inaccuracies in translations especially in the context of emotions of a particular text, leading to noisy or incorrect training data.
+These services, in our case using Google Translate\footnote{Google Cloud Translate URL: https://cloud.google.com/translate} can introduce errors or inaccuracies in translations especially in the context of emotions of a particular text, leading to noisy or incorrect training data.
 Consequently, the models may learn from erroneous information, which did negatively impact our models performance on the ED tasks.
 
+Specifically in our case of extending the original English and German datasets with translated data seems to indicate that the sentences and structure are not the same and therefore the prediction rates went down as the models got more diluted.
+
 \subsection{Findings and Implications for Improving Emotion Detection}
+Based on the improvement results in Sections \ref{sec:english-dataset-results} and \ref{sec:german-dataset-results} it seems that using translation to extend both the English and German datasets resulted in a negative outcomes.
+The best outcomes were using a simple Linear Regression model on the English and German, however based on the F1 score (EQ \ref{eq:f1-score}) of the Random Forest Classifier it seems to perform the best on both English and German.
+
+Unfortunately based on the results of extending the data our approach to also using real-time translation for multi-language models
 
@@
 \clearpage
-\section{Demonstration and Application}
-
-\subsection{Description of the developed translation application for real-time text translation}
-Details of the entire application can be found on the github repository \cite{Hoehn_Improving_Emotion_Detection_2023}.
-
-\subsection{Showcase of the application's functionality and its integration with the Emotion Detection process}
-\ldots
-
-\subsection{Demonstration of the real-time processing of translated text and emotion detection using the ML models}
-\ldots
-
-\subsection{GIT Repository Details}
-The listing the datasets used, code for the application, and details of the research project process in this research paper is openly available and hosted on a GitHub repository, allowing for easy access, reproducible, and collaborative engagement with the research findings.
-The repository is open and available under: \texttt{https://github.com/richardhoehn/mtsu.coms.qualifying-exam}\cite{Hoehn_Improving_Emotion_Detection_2023}
-\clearpage
-
 \addcontentsline{toc}{section}{References}

### 2023-08-14 paper/appendix/code.tex
Paper Updates | Savings States

@@
 \begin{minted}[fontsize=\small]{python}
-# Setup API 
+# Setup Spark Session
+from pyspark.sql import SparkSession
+
+# Import API Modules 
 from flask import Flask, request, jsonify
+from flask_cors import CORS
+import datetime
+
+# Import Translator
 from googletrans import Translator
 
-# Setup
+# Impport Tokenizer and Pipeline Tools
 from transformers import DistilBertTokenizer
 from pyspark.ml.pipeline import PipelineModel
+
+# Setup Spark Session
+sc = SparkSession \
+        .builder \
+        .master("local[*]") \
+        .appName("API-Demo") \
+        .getOrCreate()
+
+# Add Numerical Label for Emotions
+emotion_key = {
+    "boredom": 0,
+    "love": 1,
+    "relief": 2,
+    "fun": 3,
+    "hate": 4,
+    "neutral": 5,
+    "anger": 6,
+    "happiness": 7,
+    "surprise": 8,
+    "sadness": 9,
+    "worry": 10,
+    "enthusiasm": 11,
+    "empty": 12,
+    "---": 13
+}
 
@@
 app = Flask(__name__)
+CORS(app)  # Use the CORS class to enable CORS for the entire app
 translator = Translator()
@@
 def transform_data():
+    start_time = datetime.datetime.now()
     sentence_en = request.args.get('sentence_en')
+    sentence_de = request.args.get('sentence_de')
 
-    if not sentence_en:
-        return jsonify({'error': 'No Sentence (EN) provided!'})
+    print(f"English: {sentence_en}")
+    print(f"German: {sentence_de}")
 
+    if sentence_en:
         sentence_de = translator.translate(sentence_en, dest="de").text
+    elif sentence_de:
+        sentence_en = translator.translate(sentence_de, dest="en").text
+    else:
+        return jsonify({'error': 'No Sentence provided!'})
+
     print(f"English: {sentence_en}")
@@
 
+    # Get End Time
+    end_time = datetime.datetime.now()
+    elapsed_time = (end_time - start_time).total_seconds()
+
     #Build Response
@@
         "sentence": {
-            "english":sentence_en,
-            "german":sentence_de,
+            "english": sentence_en,
+            "german": sentence_de,
         },
@@
         },
+        "metadata": {
+            "spark": sc.version,
+            "datetime": {
+                "start": start_time,
+                "end": end_time,
+                "elapsedInSeconds": elapsed_time,
+            }
+        }
     }
@@
     app.run(host='127.0.0.1', port=8080)
+
 \end{minted}

### 2023-08-14 paper/main.tex
Paper Updates | Savings States

@@
 
-\usepackage{minted}
+\usepackage[newfloat]{minted}
+\usepackage{caption}
 \setminted[]{linenos, numbersep=5pt, frame=lines}
@@
 
+\newenvironment{code}{\captionsetup{type=listing}}{}
+\SetupFloatingEnvironment{listing}{name=Source Code}
+
 \usepackage{array}
@@
 \newcolumntype{M}{>{\centering\arraybackslash}m{0.85in}}
+\newcolumntype{N}{>{\centering\arraybackslash}m{0.5in}}
 \usepackage{tablefootnote}
@@
 
-\title{Improving Emotion Detection through Real-Time Translation of Text to ML Models Trained in Different Languages}
+\title{Improving Emotion Detection through Translation of Text to ML Models Trained in Different Languages}
 
@@
 \begin{abstract}
-This Qualifying Exam\footnote{PhD students of MTSU Computational and Data Science program are required to complete a Qualifying Exam in their first year in the program.} research paper investigates the potential of improving Emotion Detection (ED) through real-time translation of text to multiple Machine Learning (ML) models trained in different languages.
-Our research focused on  English and German datasets.
-This research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED analysis research and projects, essence the lack of data to better train ML models is trying to be addressed.
+This Qualifying Exam\footnote{Comp.
+and Data Science PhD students are required to complete a Qualifying Exam in their first year} research paper investigates the potential of improving Emotion Detection (ED) through translation of extended text data to multiple Machine Learning (ML) models trained in different languages.
+Our research focused on  English (X\footnote{X is formerly known as Twitter} threads) and German (voting articles) datasets to improve prediction accuracy.
+By this our research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED research and projects.
 
-By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) become larger and the hope is that better prediction rates can be achieved in ED analysis applications due to ML training having more data for learning.
-Furthermore, translating English to German to extend the German dataset and training another ML model using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates by dual processing a sentence at the same time.
+By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) becomes larger and the hope is that better prediction rates can be achieved in ED analysis applications due to ML training having more data for learning.
+Furthermore, translating English to German to extend the German dataset and training another ML model using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates even more by dual processing a sentence at the same time.
 
-In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, translated, and two ML models built that can be accessed via an API to get a prediction on an input (POST or GET) sentence both in German or English.
+In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, parsed, cleaned, translated, and two ML models built that can be accessed via an API to get a prediction by inputting sentences both in German or English via GET method using a simple query-string.
 
 Sadly the research and results point to the realization that extending the datasets by translation did not improve prediction rates of English or German.
-Some minor gains can be achieve by accessing both English and German models at the same time via the RESTful API but unfortunately the benefits may not really support the efforts.
+Some minor gains can be achieved by accessing both English and German models at the same time via the RESTful API but unfortunately the benefits may not really support the efforts.
+In conclusion it that much more data needs to be collected in order to be able to definitely answer this research question.
 \end{abstract}
@@
 
-Emotion detection has gained significant interest in recent years as a means to understand human behavior and improve communication effectiveness.
-By analyzing text, speech or facial expressions, these detection processes try to accurately identify and classify emotions expressed by individuals or groups, often times in real-time.
+Emotion detection (ED) has gained significant interest and investment\cite{Sarker2021} in recent years as a means to understand human behavior and improve communication effectiveness.
+By analyzing text, speech or facial expressions, these detection processes try to accurately identify and classify emotions expressed by individuals or groups, often in real-time.
 
-By employing a variety of ED processes, including natural language processing (NLP) and machine learning (ML), applications with the right amount of data can detect emotions with high precision.
-These ED systems find applications in diverse fields, such as sentiment analysis, customer feedback analysis, voting and news stances, and mental health monitoring, which all aim at contributing to a deeper understanding of human emotions and their impact on various aspects of life.
+By employing a variety of ED processes, including natural language processing (NLP) and machine learning (ML), applications with the right amount of data can detect emotions with high precision\cite{research-emotion-recognition-for-online-learning}.
+These ED systems find applications in diverse fields, such as sentiment analysis, customer feedback analysis, voting and news stance prediction, and mental health monitoring, which all aim at contributing to a deeper understanding of human emotions and their impact on various aspects of life.
 
 \subsection{Human Emotion as a Theoretical Framework}
+\label{sec:human-emotion-as-a-theoretical-framework}
+The concept and number of emotions can vary depending on the theoretical framework or model being considered.
+One well-known model is the Plutchik's Wheel of Emotions seen in Figure \ref{fig:plutchik-wheel}, which proposes eight (8) primary emotions and their contrasting pairs.
+In this research paper we will be focusing on the basic eight (8) emotions and not on their sub-emotions or contrasting pairs.
+This allows the ML models being trained and built to be a multi-class classification model with eight labeled options.
 
-The concept and number of emotions can vary depending on the theoretical framework or model being considered.
-One well-known model is the Plutchik's Wheel of Emotions Figure \ref{fig:plutchik-wheel}, which proposes eight (8) primary emotions and their contrasting pairs.
-In this research paper we will be focusing on the basic eight emotions and not on their sub-emotions or contrasting pairs.
-This allows the ML models being trained and built to be a multi-classification model with eight label options.
-
-Listed below are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp} that we will be focusing on both in English and German, a detailed reference of emotions used in this research can be found in Table \ref{table:dataset_labels} on page~\pageref{table:dataset_labels}.
+The following are the eight primary emotions according to the Plutchik's Wheel of Emotions \cite{Tromp} that we will be focusing on both in our English and German datasets.
+A detailed reference of emotions used in this research can be found in Table \ref{table:dataset_labels} on page~\pageref{table:dataset_labels}.
 
@@
 
-It's important to note that emotions are complex and multifaceted, and different emotion models or frameworks may propose alternative categorizations or variations in the number of primary emotions.
+It's important to note that emotions are complex and multifaceted, and different emotion models and/or frameworks may propose alternative categorizations or variations in the number of primary emotions.
 However for the scope of this research project the most common eight (8) emotions are used as described above.
-This allows the ML models being trained and built to be a multi-classification model with eight label options.
+This allows the ML models being trained and built to be a multi-class classification model with eight labeled classes.
 
-By aligning the emotional labels across the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets that are being used in this research all eight listed above were used.
-Details on the linkage of the two datasets and their counts are described in Section \ref{sec:data-procurement-process} with an emphasis on Table \ref{table:dataset_labels}'s cross reference on page~\pageref{table:dataset_labels} of German and English languages.
+By aligning the emotional labels across the English \cite{english-dataset-twitter} and German \cite{german-dataset-cheese} datasets that are being used in this research all eight listed above were linked to each other.
+Details on the linkage of the two datasets and their counts from data are described in Section \ref{sec:data-procurement-process} with an emphasis on Table \ref{table:dataset_labels}'s cross reference on page~\pageref{table:dataset_labels} of German and English language data.
 
 \subsection{Significance of Emotion Detection in Machine Learning}
-Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions (chatbots\cite{chatbot-cognitive-awareness}) to just name a few.
+Emotion Detection (ED) analysis in Machine Learning (ML) models has gained significant attention due to its potential applications in various fields, including customer sentiment analysis, stance detection in regard to a specific targets such as news and voting\cite{mascarell-etal-2021-stance}, mental health monitoring\cite{Colonnello}, and human-computer interactions such as chat-bots\cite{chatbot-cognitive-awareness} to just name a few.
 
-By accurately identifying and understanding emotions from text data, ML applications can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}.
+By accurately identifying and understanding emotions from text data, ML applications can assist in improving user experiences, decision-making processes, and overall human-machine interactions in a positive manner\cite{Colonnello, mascarell-etal-2021-stance}, with most of these interactions being processed in real-time.
 
-As an example, the primary goals of chat-bots are entertainment, social contact, and novelty interaction, with a strong emphasis on productivity, scale-ability and cost reductions.
-In chat-bots, there has been a immense requirement to simulate human-like characteristics and behavior during human-machine interactions\cite{emotion-detection-literature-review}.
+As an example, the primary goals of chat-bots are entertainment, social contact, support, and novelty interaction, with a strong emphasis on productivity, scale-ability and cost reductions.
+In the case of chat-bots, there has been a immense requirement to simulate human-like characteristics and behaviors during human-machine interactions\cite{emotion-detection-literature-review}.
 Per Kusal et al.
 \cite{chatbot-cognitive-awareness} these chat-bots also improve customer engagement by offering friendliness, comforters, flexibility, and efficient assistance.
-Chat-bots should provide customers with more engaging responses, directly addressing their issues or questions.
-The goal of deploying ED applications is for users perceive chat-bots as companions rather than simple assistants, which in turn helps the overall process of getting to a positive result.
+As such chat-bots should provide customers with more engaging responses, directly addressing their issues or questions.
+The goal of deploying ED to these applications is for users perceive chat-bots as companions rather than simple assistants, which in turn helps the overall process of getting to a positive result.
 Per Adikari et al.
-the majority of user requests are emotional rather than informative; therefore the need that chat-bots gain the capacity to respond emotionally to customers due to machine learning and sentiment analysis evolution\cite{chatbot-cognitive-awareness} is very important and yields for many applications that employee ED a way to improve interactions with machines.
+the majority of user requests are emotional rather than informative; therefore the need for chat-bots to gain the capacity to respond emotionally to customers via machine learning and sentiment analysis evolution\cite{chatbot-cognitive-awareness} is very important and yields for many applications that use ED a way to improve interactions with machines.
 
@@
 Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions.
-This in turn contributes to the advancement of artificial intelligence, natural language processing (NLP) and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions} to go about our daily lives.
+Our research in turn contributes to the advancement of artificial intelligence, natural language processing (NLP) and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions} to go about our daily lives.
 
-Lastly, in Chen et al.\cite{research-emotion-recognition-for-online-learning} paper based on emotion detection for online learning, it states interestingly that with emotion recognition research, we need to focus on accuracy and real-time performance in order to apply ED based on physiological signals to solve practical problems; the key part of their claim is that in order for ED to be practical, real-time methods needs to be used and deployed for maximum results.
+Lastly, in the paper by Chen et al.\cite{research-emotion-recognition-for-online-learning} based on emotion detection for online learning it states, interestingly, that with emotion recognition research, we need to focus on accuracy and real-time performance in order to apply ED based on physiological signals to solve practical problems; the key part of their claim is that in order for ED to be practical, real-time methods need to be used and deployed for maximum results.
 
 \subsection{Challenges: Lack of Labeled Datasets and Language Fragmentation}
-Emotion Detection is an immensely growing field, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}.
+Emotion Detection is a fast growing field\cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}.
 The primary reason of the lack of data stems from the face that emotion detection data requires primarily supervised learning data.
@@
 The market for emotion recognition and detection software is forecast to reach at least \$3.8 billion by 2025 \cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}.
-And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, customer service, news and stance analysis, and a multiple of others, it is fitting to work on research to solve and / or improve issues in regards to prediction rates and lack of teaching / learning data.
+And since ED is already being incorporated into products for purposes such as marketing, robotics, driver safety, customer service, news and stance analysis, and a multiple of others, it is fitting to work on research to solve and / or improve issues in regards to prediction rates due to lack of data and language fragmentation.
 
-With the before mentioned issues regarding the quantity (lack of) and fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED.
+With the before mentioned issues regarding the quantity (lack of) and language fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED.
 The aim of the research is that by use of translation that the predictability of ED can be improved by extending an English dataset with translations from German and also process look-ups in real-time language conversion and processing.
@@
 
-The research project aims at answering the following research questions:
+The primary objective of this Qualifying Exam research project is to address three (3) research questions, each aimed at improving the accuracy of predictive ED models in the context of English and German language datasets.
 
 \begin{itemize}
-\item Can by translating English data to German and extending an original German dataset increase the predictability of German ED model?
-\item Can by translating German data to English and extending an original English dataset increase the predictability of English ED model?
-\item Can by translating in real-time an input to multiple languages improve the predictability based on the combined output of two models that were trained in English and German to better the prediction results.
+\item The first hypothesis is that by introducing English data translated into German and extending an original German dataset, will the added text lead to an enhancement in its predictive capabilities.
+(Section \ref{sec:analysis-of-impact-of-extending-datasets})
+\item Similar to the first, can by translating German data to English and extending an original English dataset increase the predictability of English ED model?
+(Section \ref{sec:analysis-of-impact-of-extending-datasets})
+\item The third research question shifts the focus to real-time translation and its impact on prediction.
+Can by translating in real-time an input to multiple languages improve the predictability based on the combined output of two models that were trained in English and German to better the prediction results.
+(Section \ref{sec:impact-using-api-for-real-time-translation})
 \end{itemize}
 
-With these three (3) objectives our approach will be to procure, translate, train, and evaluate benefits of extending datasets by translation to improve emotion detection.
+In summary, this research project centers around investigating innovative ways to enhance the predictability of Emotion Detection models in both English and German.
+With these three (3) objectives from above our methodology will be to procure, translate, train, and evaluate benefits of extending datasets by translation to improve emotion detection.
 
@@
 \subsection{Emotion Detection Importance and Diverse Populations}
-Emotion Detection (ED) holds significant importance in various domains, including psychology, social sciences, and human-computer interaction.
-It enables the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states both in real-time but also as a type barometer for measuring sentiments from the past versus the current time.
+Emotion Detection (ED) holds significant importance\cite{Sarker2021} in various domains, including psychology, social sciences, and human-computer interaction.
+By use of ML models\cite{CHOWANDA-2021821} the analysis of human emotions at scale, providing valuable insights into individual and collective emotional states both in real-time but also as a type barometer for measuring sentiments from the past versus the current time.
 In Chowanda et al.
@@
 
-\subsection{Data procurement process of ED datasets in English and German}
+\subsection{Data Procurement and Parsing of ED datasets in English and German}
 \label{sec:data-procurement-process}
-The data procurement was relatively straight forward and in essence entailed using Google search terms for the Emotion Detection in English and German text languages.
-Some of the main German data was obtained from the dataset built by ETH's Emotion and Stance Detection for German Text \cite{mascarell-etal-2021-stance} which also hosts a good website with their findings \footnote{https://mtc.ethz.ch/research/natural-language-processing/emotion-stance.html}.
-The English dataset was downloaded from Kaggle \footnote{https://www.kaggle.com/datasets/pashupatigupta/emotion-detection-from-text} based on Tweets collected in 2021.
-The English dataset contains over 38,000 tweets and their corresponding emotion labeled.
-Unfortunately the German dataset only contains over 2,000 sentences and their respective emotions associated.
-this imbalance of the two dataset sizes will be corrected in the subsequent sections.
+The data procurement was relatively straight forward and once found by using multiple Google search terms for the Emotion Detection in English and German text languages.
+Some of the main German data was obtained from the dataset built by ETH's Emotion and Stance Detection for German Text \cite{mascarell-etal-2021-stance} which also hosts a good website with their findings\footnote{https://mtc.ethz.ch/research/natural-language-processing/emotion-stance.html}.
+The English dataset was downloaded from Kaggle\footnote{https://www.kaggle.com/datasets/pashupatigupta/emotion-detection-from-text} based on Tweets collected in 2021.
+The English dataset contains over 38,000 tweets and their corresponding emotion's labeled.
+Unfortunately the German dataset only contained just over 2,500 sentences and their respective emotions associated with it.
+Unfortunately this imbalance (see Table \ref{table:file-details}) of the two dataset sizes will had to be corrected and is explained in detail in the subsequent sections.
 
@@
 
-An important part of the process is to make sure we match the English emotions to those of the German language emotions.
+An important part of the process is to make sure we match / link the English emotions to those of the German language emotions.
 Since the two languages are dissimilar we opted to match exact emotions to each other that can be seen in Table \ref{table:dataset_labels} below.
+The linking yielded our eight (8) emotions we previously discussed in Sections \ref{sec:human-emotion-as-a-theoretical-framework}.
+Details on how we formatted the the two (English and German) datasets to each other can be viewed in the Appendix \ref{appendix:dataset_english} and \ref{appendix:dataset_german}.
 
@@
 
-Details on how we formatted the the two (English and German) datasets to each other can be viewed in the Appendix \ref{appendix:dataset_english} and \ref{appendix:dataset_german}
+The processing of two ED datasets to be synchronized or linked to the emotion is important.
+The first will be in English and the second in German.
+Each dataset will be translated into the other's language for extension purposes.
 
-The processing of two ED datasets to be synchronized to the emotion is important.
-The first will be in English and the second in German.
-Each dataset will be translated into the other’s language for extending purposes.
-And each row will hold the English and German sentence.
-As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German dataset.
+We decided to build the data to support that each row of each dataset will hold the English and German sentence.
+As an example Figure \ref{fig:dataset-extension} depicts the original English dataset will be extended by the translated German (German $\Rightarrow$ English) dataset.
 By this simple process the English dataset got larger for training purposes.
@@
 
-Once the two datasets have been translated, we now extend the original language dataset with the translated one.By this approach we are in essence extending the original datasets to create a larger one with the foreign language data.
+Once the two datasets have been translated, we now extend the original language dataset with the translated one.
+By this approach we are in essence extending the original datasets to create a larger one with the foreign language data.
 With both extended datasets we can now train our ML models for English and German.
 
-\subsection{Development of a translation application for converting different languages into English and vice versa}
+\subsection{Translation Application for Converting English to German and Vice Versa}
 Translation of the emotion datasets was performed with Google Translator based on the Medium article\cite{Nidhaloff_how_to_translate_text_with_python} in which the author describes options for simple API translations.
@@
 
+\begin{code}
+\centering
 \begin{quote}
 \begin{minted}[samepage, fontsize=\small]{python}
-#
-# Using Deep Translator to leveage Google Translate
+# Using Deep Translator to leverage Google Translate
 # Link: https://cloud.google.com/translate/docs/reference/libraries/v2/python
@@
 print(translated_back) # Chocolate milk tastes much better through a straw.
+\end{minted}
+\captionof{listing}{Google Translate Example Code}
+\end{quote}
+\end{code}
 
-\end{minted}
-\end{quote}
-
-Based on the above translation example the CSV (English) and JSON (German) datasets were individually loaded, parsed, and translated to their respective counter-part languages.
+Based on the above translation example the CSV\footnote{CSV = Comma Separated Values} (English) and JSON\footnote{JSON = JavaScript Object Notation} (German) datasets were individually loaded, parsed, and translated to their respective counter-part languages.
 Details of the application can be found in the Appendix \ref{appendix:dataset_english} for English and Appendix \ref{appendix:dataset_german} for the German processing.
@@
 \subsection{Training Multiple ML models with PySpark}
-After the completion of translating and processing the Englsih and German datasets we come to training our models.
+After the completion of translating and processing the English and German datasets we come to training our models.
 Using the datasets from Tables \ref{table:pd_de_translated}, \ref{table:pd_en_translated}, and \ref{table:ml_models} that list an approximate $\sim$1,275 and $\sim$2,775 features for training.
@@
 
+\begin{code}
+\centering
 \begin{quote}
 \begin{minted}[samepage, fontsize=\small]{python}
-#
 # Split the English and German Dataframes for Training and Testing
@@
 df_de_train_extended = df_de_train.union(df_en.select(*df_de_train.columns))
-
 \end{minted}
 \end{quote}
+\captionof{listing}{Linking the English and German Emotions to the same label}
+\end{code}
 
@@
 \end{tabular}
-\caption{ML Models and their respective settings}
+\caption{ML Models and their respective Train and Test Rows}
 \label{table:ml_models}
@@
 We grouped the label by count from the test data (both English and German) and compared the predicted result counts to those.
-It became very clear that there was a strong bias on the GLM (Linear Regression) model that essentially proposed that all sentences evalued to the same output both in English and German.
+It became very clear that there was a strong bias on the GLM (Linear Regression) model that essentially proposed that all sentences evaluated to the same output both in English and German.
+
+\subsection{Creating an API for Translation and Parallel Processing}
+\label{sec:create-api-for-translation-and-parallel-processing}
+
+In order for a user to access and get prediction results from all three (Random Forest, Naives Bayes, and Linear Regression) ML models trained in both English and German in real-time we crated a Jupyter Notebook that starts an Apache PySpark\footnote{PySpark: https://spark.apache.org/docs/latest/api/python/index.html} session that is accessed via the local server listen on \texttt{https://127.0.0.1:8080}.
+The details of the RESTful the server can be found in Appendix \ref{appendix:code-api-server}.
+
+The server listens for GET requests and evaluates a querystring either called \texttt{sentence\_en} or \texttt{sentence\_de} if an English sentence is sent then the server translates to German and if a German is sent then it translates to English.
+Each sentence is then tokenized, formatted, and processed on the previously trained ML models in both languages.
+
+Once all the the ML model predictions have completed, we then lookup the emotion associated with the individually predicted labels and build a response \texttt{JSON} object to pass back the calling application.
+The \texttt{JSON} object (Source Code: \ref{code:json-object}) contains the original sentence and translated, plus each emotions predicted label by the six previously trained models.
+In addition the main data, and \texttt{metadata} object is created that holds the PySpark version, start, and end times, including the processing time in seconds.
+
+\begin{code}
+\begin{quote}
+\begin{minted}[samepage, fontsize=\small]{json}
+{
+  "metadata": {
+    "datetime": {
+      "elapsed": 1.651038,
+      "end": "Mon, 14 Aug 2023 12:10:35 GMT",
+      "start": "Mon, 14 Aug 2023 12:10:33 GMT"
+    },
+    "spark": "3.3.2"
+  },
+  "predictions": {
+    "de_lr": "enthusiasm",
+    "de_nb": "relief",
+    "de_rfc": "enthusiasm",
+    "en_lr": "neutral",
+    "en_nb": "love",
+    "en_rfc": "happiness"
+  },
+  "sentence": {
+    "english": "What a wonderful world this is!",
+    "german": "Was für eine wundervolle Welt das ist!"
+  }
+}
+\end{minted}
+\captionof{listing}{JSON output from API}
+\label{code:json-object}
+\end{quote}
+\end{code}
+
+\subsection{Evaluation Metrics and Methodologies for Measuring the Performance}
+\label{sec:evaluation-metrics-and-methodologies}
+The evaluation of ML models trained on the English, German and their respective extension datasets is a critical step in assessing their performance and reliability.
+We chose to use the MulticlassClassificationEvaluator\footnote{URL: https://spark.apache.org/docs/latest/api/python/reference/api/pyspark.ml.evaluation.MulticlassClassificationEvaluator.html} because is a well known tool in the field of natural language processing (NLP) research, specifically in multi-class classification tasks such as emotion detection that has multiple labels associated with it; in our case eight (8) see Table \ref{table:dataset_labels} on Page \pageref{table:dataset_labels}.
+Generally this evaluator is used to evaluate the performance of machine learning (ML) models, using transformer-based architectures like BERT, RoBERTa, or DistilBERT\cite{Sanh2019-DistilBERT-AD}, with DistilBERT\cite{Sanh2019-DistilBERT-AD} being the one we used in this research project.
 
@@
 
-In order for each of the three ML models to be processed at the same time we crated a Jupyter Notebook that starts an Apache PySpark\footnote{PySpark: https://spark.apache.org/docs/latest/api/python/index.html} session that is accessed via the local server running that accepts RESTful API calls with details of the server in Appendix \ref{appendix:code-api-server}.
-
-\subsection{Evaluation Metrics and Methodologies for Measuring the Performance}
-\label{sec:evaluation-metrics-and-methodologies}
-The evaluation of ML models trained on the English, German and their respective extension datasets is a critical step in assessing their performance and reliability.
-We chose to use the MulticlassClassificationEvaluator\footnote{URL: https://spark.apache.org/docs/latest/api/python/reference/api/pyspark.ml.evaluation.MulticlassClassificationEvaluator.html} because is a well known tool in the field of natural language processing (NLP) research, specifically in multi-class classification tasks such as emotion detection that has multiple labels associated with it; in our case eight (8) see Table \ref{table:dataset_labels} on Page \pageref{table:dataset_labels}.
-Generally this evaluator is used to evaluate the performance of machine learning (ML) models, using transformer-based architectures like BERT, RoBERTa, or DistilBERT\cite{Sanh2019-DistilBERT-AD}, with DistilBERT\cite{Sanh2019-DistilBERT-AD} being the one we used in this research project.
-
 The MulticlassClassificationEvaluator provides the essential evaluation metrics, including accuracy, Precision (eq: \ref{eq:precision}), Recall (eq: \ref{eq:recall}), and F1 Score (eq: \ref{eq:f1-score}) for ED classifications.
@@
 As an added example simply using the accuracy to evaluate a model can be miss-leading and therefore in our case using the F1 Score, that is directly correlated to the Precision (eq: \ref{eq:precision} and Recall (eq: \ref{eq:recall}) is immensely important to understanding the ones models and finding the best of for the task of Emotion Detection using Original and Extended datasets.
+
+\subsection{Using an API to Translate in Real-Time for Dual Prediction}
+\label{sec:using-api-to-translate-in-ral-time}
+The last stage of this research project was devoted to testing the API by use of a simple webpage.
+In order to accurately test and get real-world results and a sense on the real-time processing of a Emotion Detection a simple webpage was build to give a user the sense of the conclusion of the project.
+In Figure \ref{fig:demo-api-webpage} a brief screenshot of some testing can be seen.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.85\textwidth]{images/Demo-Api-2.png}
+    \caption{Demo API Webpage}
+    \label{fig:demo-api-webpage}
+\end{figure}
+
+In the screenshot (Figure: \ref{fig:demo-api-webpage}) example a sentence was entered in the English text-box and the "Predict" button pressed.
+The API took about $\sim$1.6seconds to translate the English to German and then process both sentences on the pre-trained ML models, build the \texttt{JSON} object and return it to the application.
 
@@
 
+\subsection{Impact Using API for Real-Time Translation}
+\label{sec:impact-using-api-for-real-time-translation}
+With the help of the webpage (Figure \ref{fig:demo-api-webpage} on \pageref{fig:demo-api-webpage}) a sense of the inaccuracy was also presented.
+Each round trip to the API took an between 1.4 and 1.6seconds.
+The prediction results mirrored those of Tables \ref{table:en_testing_results} and \ref{table:de_testing_results} which in turn made the usefulness of the translation not practical as can be seen in Table \ref{table:api-test-results} below.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | N | M | L | N | L | N | }
+    \hline
+    \multicolumn{6}{|c|}{Webpage API Test} \\
+    \hline
+     & 
+    \multicolumn{2}{c|}{English} &
+    \multicolumn{2}{c|}{German} & 
+     \\
+    \hline
+    Test &
+    Emotion & Sentence &
+    Emotion & Sentence &
+    Time \\
+    \hline
+    A &
+    Worry & This is bad idea and we need to stop right now &
+    Anger & Dies ist eine schlechte Idee und wir müssen jetzt aufhören &
+    1.49s \\
+    \hline
+    B &
+    Happiness & What a wonderful world this is! &
+    Relief & Was fur eine wundervolle Welt das ist! &
+    1.65s \\
+    \hline
+    C &
+    Happiness & I am so happy about this news! &
+    Fun & Ich freue mich sehr uber diese Nauigkeiten! &
+    1.64s \\
+    \hline
+    D &
+    Sadness & Gosh I hate doing this today. &
+    Relief & Meine Güte, ich hasse es heute. &
+    2.56s \\
+    \hline
+\end{tabular}
+\caption{API Webpage Test Results}
+\label{table:api-test-results}
+\end{table}
+
 \subsection{Findings and Implications for Improving Emotion Detection}

### 2023-08-15 paper/main.tex
Working on Paper

@@
 and Data Science PhD students are required to complete a Qualifying Exam in their first year} research paper investigates the potential of improving Emotion Detection (ED) through translation of extended text data to multiple Machine Learning (ML) models trained in different languages.
-Our research focused on  English (X\footnote{X is formerly known as Twitter} threads) and German (voting articles) datasets to improve prediction accuracy.
+Our research focused on  English and German text data to improve prediction accuracy.
 By this our research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED research and projects.
 
-By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) becomes larger and the hope is that better prediction rates can be achieved in ED analysis applications due to ML training having more data for learning.
+By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) becomes larger and the hope is that better prediction rates can be achieved in ED analysis applications due to more ML training data.
 Furthermore, translating English to German to extend the German dataset and training another ML model using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates even more by dual processing a sentence at the same time.
 
-In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, parsed, cleaned, translated, and two ML models built that can be accessed via an API to get a prediction by inputting sentences both in German or English via GET method using a simple query-string.
+In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and with this paper datasets both in English and German were collected, parsed, cleaned, translated, and two ML models built that can be accessed via an API to get a prediction by inputting sentences both in German or English via GET method using a simple query-string.
 
-Sadly the research and results point to the realization that extending the datasets by translation did not improve prediction rates of English or German.
-Some minor gains can be achieved by accessing both English and German models at the same time via the RESTful API but unfortunately the benefits may not really support the efforts.
-In conclusion it that much more data needs to be collected in order to be able to definitely answer this research question.
+Sadly the results point indicate that extending the datasets by translation did not improve prediction rates of English or German.
+Some minor gains may be achieved by accessing both English and German models at the same time via the RESTful API but unfortunately the benefits may not really support the efforts.
+We believe that in review that due to the many classes in this multi-class classification model that overfitting became an issue for certain labels and created a type memorization rather than learning effect on the data.
+In conclusion much more data needs to be collected or by use of AI created in order to be able to definitely answer this research question.
 \end{abstract}
@@
 Based on the improvement results in Sections \ref{sec:english-dataset-results} and \ref{sec:german-dataset-results} it seems that using translation to extend both the English and German datasets resulted in a negative outcomes.
-The best outcomes were using a simple Linear Regression model on the English and German, however based on the F1 score (EQ \ref{eq:f1-score}) of the Random Forest Classifier it seems to perform the best on both English and German.
+The best outcomes were using a simple Linear Regression model on the English and German, however based on the F1 Score (Eq: \ref{eq:f1-score}) of the Random Forest Classifier it seems to perform the best on both English and German.
+This is because the false positive amount of one score on the accuracy is not taken into account and can easily skew the results when using a multi-class classification model.
 
-Unfortunately based on the results of extending the data our approach to also using real-time translation for multi-language models
+Unfortunately based on the results of extending the data our approach to also using real-time translation for multi-language models by use of creating an API and translating text in real-time for both German and English seems to be unsatisfactory for any meaningful improvement.
+
+With both these findings, we believe that the data of only using $\sim$1,500 (original dataset) and $\sim$2,700 (extended dataset) to train models is insufficient\cite{overfitting-overview} when using a multi-class classification problem.
+By inference, if we have only 1,500 and divide (eq: \ref{eq:sentence-per-class}) those by our eight (Table: \ref{table:dataset_labels}) classes we simply do not have enough data to truly train the ML on the classes.
+
+\begin{equation}
+\begin{split}
+\frac{Features}{Labels}= \frac{Sentences}{Classes}=\frac{1,500}{8}=185 \\
+or=\frac{2,700}{8}=337
+\end{split}
+\label{eq:sentence-per-class}
+\end{equation}
+
+Taking this this point that the data is insufficient the implication of lack of classified data can result in a few major issues for our ML models.
+One, the lack of context for the models to work properly and learn intricate details of the English and German language.
+And second, in some cases where lack of class data is present the models memorize existing sentences and words rather than learning underlying patterns of said language.
+Both these issues are related to an phenom called overfitting\cite{overfitting-overview} present in many regression problems.
+
+Options for improvement are on the simple side to get our hands on more data, the second is much more nuanced, in that we find ways to take existing and labeled sentences and have machine find derivatives of such so we can create more intricate patterns for the machines to learn from.
 
@@
 It's crucial to ensure the quality and accuracy of extended data, especially when language translation is involved, to prevent introducing more noise than meaningful information into the training process.
+The implications of lack of data in ED analysis presented it self in this research that it wanted to try and overcome.
 Although the research did not garner any significant improvements by translating the German to English as a means on extending the English dataset, it actually made it the prediction results worse.
+We believe one of the main culprits of this that the data for learning based on eight classes was insufficient.
+As can be seen in Equation \ref{eq:sentence-per-class} from the previous section, the models most likely started memorizing the sentence rather than learn the intricacies of the labeled classes.
+This conclusion is arrived because some labels worked better than other, those that worked better were also more plentiful available in the training data.
 
-Future work includes testing and translating different models by using different translation services.
+There tow areas of future research; one is to further testing and translating different models by using different translation services.
 Using Google Cloud's translation service was adequate for this research, however there are certainly better translation services that are available but are based on paid subscriptions.
+Second, another path to extend data would be to have AIs convert sentences to be similar but not the same by passing the sentence and the labeled class to an AI to generate more sentences with similar emotion attached to them.
+The hope would be that by mushrooming sentences that are based on an emotion and pass the sentence that we can create more labeled data.
 

### 2023-08-16 paper/main.tex
Paper Updates - Feedback Review

--- 2023-08-14 paper/main.tex
@@
 This Qualifying Exam\footnote{Comp.
-and Data Science PhD students are required to complete a Qualifying Exam in their first year} research paper investigates the potential of improving Emotion Detection (ED) through translation of extended text data to multiple Machine Learning (ML) models trained in different languages.
-Our research focused on  English (X\footnote{X is formerly known as Twitter} threads) and German (voting articles) datasets to improve prediction accuracy.
-By this our research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED research and projects.
+and Data Science PhD students are required to complete a Qualifying Exam in their first year} research paper investigates enhancing Emotion Detection (ED) by translating extended text data into various Machine Learning (ML) models trained in distinct languages.
+We focused on English and German text data to enhance prediction accuracy, aiming to overcome challenges arising from limited labeled datasets and language fragmentation in ED research.
 
-By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) becomes larger and the hope is that better prediction rates can be achieved in ED analysis applications due to ML training having more data for learning.
-Furthermore, translating English to German to extend the German dataset and training another ML model using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates even more by dual processing a sentence at the same time.
+Expanding an original English dataset with translated German data increases the training data's volume, potentially improving prediction rates in ED applications.
+Additionally, translating English to German to extend the German dataset and accessing in real-time ML models trained in both could further improve prediction rates.
 
-In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, parsed, cleaned, translated, and two ML models built that can be accessed via an API to get a prediction by inputting sentences both in German or English via GET method using a simple query-string.
+For presentation purposes, datasets in both English and German were collected, parsed, cleaned, and translated.
+Multiple ML models trained in English and German where built, and made accessible via an API for predictions in either language including real-time translation using a GET method.
 
-Sadly the research and results point to the realization that extending the datasets by translation did not improve prediction rates of English or German.
-Some minor gains can be achieved by accessing both English and German models at the same time via the RESTful API but unfortunately the benefits may not really support the efforts.
-In conclusion it that much more data needs to be collected in order to be able to definitely answer this research question.
+The research findings suggest that the extension of datasets through translation has not yielded improvements in predictive accuracy for both English and German languages.
+Modest enhancements could potentially be achieved by concurrently accessing English and German models through real-time translation via the RESTful API; however, the benefits may not fully justify the efforts.
+
+We posit that the prevalence of numerous classes in this multi-class classification model has contributed to instances of overfitting across several labels/classes.
+This occurrence, in turn, has led to a memorization effect rather than facilitating genuine learning within the models.
+
+In conclusion, it becomes evident that a more substantial accumulation of data is required, or an innovative approach involving the utilization of AI to generate analogous data must be employed to comprehensively address this research question.
 \end{abstract}
@@
 
-Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions.
+Due to the above mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions.
 Our research in turn contributes to the advancement of artificial intelligence, natural language processing (NLP) and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions} to go about our daily lives.
@@
 The primary reason of the lack of data stems from the face that emotion detection data requires primarily supervised learning data.
-In in turn requires time and effort to construct, catalog, and procure,; this all is mostly created and labeled by humans.
+In turn requires time and effort to construct, catalog, and procure,; this all is mostly created and labeled by humans.
 To compound on the issue of the lack of labeled data; the many datasets that are available are in many cases in multiple languages - not all are in English since emotions that are linked to text are contextual in nature.
@@
 
-With the before mentioned issues regarding the quantity (lack of) and language fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED.
+With the above mentioned issues regarding the quantity (lack of) and language fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED.
 The aim of the research is that by use of translation that the predictability of ED can be improved by extending an English dataset with translations from German and also process look-ups in real-time language conversion and processing.
@@
 This provided some help, but quickly established itself as time consuming task and need of expertise in the subject matters that could not outweigh the results.
-Overcoming the scarcity of labeled datasets and embracing innovative approaches like the before mentioned study by Kapoor\cite{KAPOOR2023120882} et al.
+Overcoming the scarcity of labeled datasets and embracing innovative approaches like the above mentioned study by Kapoor\cite{KAPOOR2023120882} et al.
 and their analysis can contribute to the advancement of ED research.
@@
 
-It is important to note that in order to accurately test and compare the original and the extended datasets to each other we made sure that the test data was the same.
-to do this we randomly extracted 15\% of the original datasets (English and German) before extending the original with the translated datasets.
+It is important to note that in order to accurately test and compare the original and the extended datasets to each other we made sure that the test data was the same on each case.
+To do this we randomly extracted 15\% of the original datasets (English and German) and saved it as a test dataset prior to extending the original with the translated sentence data.
 
@@
 Based on the improvement results in Sections \ref{sec:english-dataset-results} and \ref{sec:german-dataset-results} it seems that using translation to extend both the English and German datasets resulted in a negative outcomes.
-The best outcomes were using a simple Linear Regression model on the English and German, however based on the F1 score (EQ \ref{eq:f1-score}) of the Random Forest Classifier it seems to perform the best on both English and German.
+The best outcomes were using a simple Linear Regression model on the English and German, however based on the F1 Score (Eq: \ref{eq:f1-score}) of the Random Forest Classifier it seems to perform the best on both English and German.
+This is because the false positive amount of one score on the accuracy is not taken into account and can easily skew the results when using a multi-class classification model.
 
-Unfortunately based on the results of extending the data our approach to also using real-time translation for multi-language models
+Considering the outcomes of data extension via translation, our approach of utilizing real-time translation for multi-language models through the creation of an API and translating text in real-time for both German and English appears to be unsatisfactory in terms of achieving any substantial improvement as well.
+
+With both these findings, we believe that the data of only using $\sim$1,500 (original dataset) and $\sim$2,700 (extended dataset) to train models is insufficient\cite{overfitting-overview} when using a multi-class classification problem.
+By inference, if we have only 1,500 and divide (eq: \ref{eq:sentence-per-class}) those by our eight (Table: \ref{table:dataset_labels}) classes we simply do not have enough data to truly train the ML on the classes.
+
+\begin{equation}
+\begin{split}
+\frac{Features}{Labels}= \frac{Sentences}{Classes}=\frac{1,500}{8}=185 \\
+or=\frac{2,700}{8}=337
+\end{split}
+\label{eq:sentence-per-class}
+\end{equation}
+
+Taking into account the insufficiency of data, the implication of a lack of classified data can lead to several significant issues for our ML models.
+Firstly, there's an absence of context for the models to operate effectively and grasp the intricate nuances of both the English and German languages.
+Secondly, in cases where class data is lacking, the models tend to memorize existing sentences and words, rather than learning the underlying patterns of the language.
+Both of these issues are connected to a phenomenon called overfitting\cite{overfitting-overview}, which is prevalent in many regression problems.
+
+Options for improvement range from the straightforward approach of acquiring more data to the more nuanced strategy of leveraging existing labeled sentences and employing machines to generate derivatives.
+This would facilitate the creation of more intricate patterns for the machines to learn from.
 
@@
 It's crucial to ensure the quality and accuracy of extended data, especially when language translation is involved, to prevent introducing more noise than meaningful information into the training process.
+The implications of lack of data in ED analysis presented it self in this research that it wanted to try and overcome.
 Although the research did not garner any significant improvements by translating the German to English as a means on extending the English dataset, it actually made it the prediction results worse.
+We believe one of the main culprits of this that the data for learning based on eight classes was insufficient.
+As can be seen in Equation \ref{eq:sentence-per-class} from the previous section, the models seemingly resorted to memorizing the sentence rather than learning the intricacies of the labeled classes.
+This inference is drawn that certain labels exhibited better F1 Score performance, primarily due to their greater availability in the training data.
 
-Future work includes testing and translating different models by using different translation services.
-Using Google Cloud's translation service was adequate for this research, however there are certainly better translation services that are available but are based on paid subscriptions.
+Two areas of prospective research emerge: firstly, conducting further tests involving the translation of different models using diverse translation services.
+While Google Cloud's translation service sufficed for this research, superior paid subscription-based translation services are available.
+With the help of more data and better translation more concrete answers to prediction improvement may be achieved.
+Secondly, another avenue to overcome the lack of data involves employing AI to generate sentences that are similar yet distinct from the original.
+This could be achieved by providing a sentence and its associated labeled class to the AI for data creation.
+This approach aims to produce additional sentences imbued with similar emotional attributes, potentially expanding the labeled dataset.
 

### 2023-08-17 paper/main.tex
Final Paper Edits and Submission Ready

@@
 \label{sec:results-and-analysis}
-The following section lists the prediction rates of training and testing of the original and extended datasets both in English and German.
-Overall results seem to indicate that there was no measurable benefit and even negative outcomes based on adding the extended data for Emotion Detection.
-In the below sub sections details on the results are listed by the common standards of using the below equations \ref{eq:accuracy}, \ref{eq:precision}, \ref{eq:recall}, and \ref{eq:f1-score} for Accuracy, Precision, Recall, and the F1 Score.
-It should be noted that ED is a Multi-class classification problem, which means that the machine learning process assigns the predictions from the sentence data to one of several predefined emotion categories (see Table \ref{table:dataset_labels}).
+This section lists the prediction rates of training and testing of the original and extended datasets both in English and German.
+Overall results seem to indicate that there was no measurable benefit and even negative outcomes from adding the extended data for Emotion Detection.
+In the subsections details on the results are listed by the common standards of using the below equations \ref{eq:accuracy}, \ref{eq:precision}, \ref{eq:recall}, and \ref{eq:f1-score} for Accuracy, Precision, Recall, and the F1 Score.
+It should be noted that ED is a multi-class classification problem, which means that the machine learning process assigns the predictions from the sentence data to one of several predefined emotion categories (see Table \ref{table:dataset_labels}).
 
@@
 Oddly, the prediction results with the extended data where lower than using the original (non-extended) data for training.
-In the analysis and conclusions section \ref{sec:analysis-of-impact-of-extending-datasets} below details on why this may be the case are discussed in detail.
+In the Analysis and Conclusions section \ref{sec:analysis-of-impact-of-extending-datasets} details on why this may be the case are discussed in detail.
 
@@
 The accuracy (eq: \ref{eq:accuracy}) of any machine learning model is a clear and easily comprehend-able way to understand the performance of a model.
-However, it is important to consider the context and potential trade-offs with other metrics, as high accuracy alone might not capture the full picture of a model's performance.
+However, it is important to consider the context and potential trade-offs with other metrics, as high accuracy alone might not adequately capture the full picture of a model's performance.
+This can be especially true for multi-class classification models such as the ED analysis that we are working on.
+
 \begin{equation}
@@
 The Precision (eq: \ref{eq:precision}) focuses in detail on the accuracy of positive predictions (TP)\footnote{Positive Predictions = TP (True Positive)} made by the model.
-Per Google Cloud's definition the precision answers: What proportion of positive predictions where actually correct?
+According to Google Cloud's definition, precision answers the question: What proportion of positive predictions were actually correct?
+
 \begin{equation}
@@
 \end{equation}
-Since our project is a multi-class classification project we will need to actually use Micro Averaging for the Precision.
-With this in mind we extend our simple equation \ref{eq:precision} of precision to incorporate micro averaging is illustrated in equation \ref{eq:precision_micro_avg}.
+
+Since our project is a multi-class classification project we will need to utilize Micro Averaging for the Precision.
+With this in mind, we extend our simple equation \ref{eq:precision} for precision to incorporate micro averaging as defined in equation
+\ref{eq:precision_micro_avg}.
+
 \begin{equation}
@@
 Since our project is a multi-class classification project we will need to actually use Micro Averaging for the Recall in a similar fashion to how we used it on Precision.
-With this in mind we extend our simple equation \ref{eq:recall} of recall to incorporate micro averaging is illustrated in equation \ref{eq:recall_micro_avg}.
+With this in mind we extend our simple equation \ref{eq:recall} of recall to incorporate micro averaging as defined in equation \ref{eq:recall_micro_avg}.
 
@@
 
-The F1 score (eq: \ref{eq:f1-score}) in our project is important to assess the balance between precision and recall.
-In addition, the F1 score is especially relevant in multi-class classification scenarios where there are more than two classes.
+The F1 Score (eq: \ref{eq:f1-score}) in our project is important to assess the balance between precision and recall.
+In addition, the F1 Score holds significance in multi-class classification scenarios where there are more than two classes.
 
@@
 
-Lastly, we used micro averaging for the Recall, Precision, and F1 Score due to fact it give each label (Emotion see Table \ref{table:dataset_labels}) equal weight regardless of the class label and the number of labels counted.
+Lastly, we employed micro averaging for the Recall, Precision, and F1 Score due to the fact that it assigns equal weight to each label (Emotion, as seen in Table \ref{table:dataset_labels}), regardless of the class label and the number of labels taken into account.
+This approach ensures that the evaluation metrics give uniform importance to all labels, making it particularly suitable for scenarios where class imbalances exist or where an fair assessment of model performance across all emotion classes is required.
 
@@
 \subsection{Findings and Implications for Improving Emotion Detection}
+\label{sec:findings-and-implications-for-improving}
 Based on the improvement results in Sections \ref{sec:english-dataset-results} and \ref{sec:german-dataset-results} it seems that using translation to extend both the English and German datasets resulted in a negative outcomes.
@@
 It's crucial to ensure the quality and accuracy of extended data, especially when language translation is involved, to prevent introducing more noise than meaningful information into the training process.
-The implications of lack of data in ED analysis presented it self in this research that it wanted to try and overcome.
-Although the research did not garner any significant improvements by translating the German to English as a means on extending the English dataset, it actually made it the prediction results worse.
-We believe one of the main culprits of this that the data for learning based on eight classes was insufficient.
-As can be seen in Equation \ref{eq:sentence-per-class} from the previous section, the models seemingly resorted to memorizing the sentence rather than learning the intricacies of the labeled classes.
-This inference is drawn that certain labels exhibited better F1 Score performance, primarily due to their greater availability in the training data.
+The implications of the lack of data in ED analysis presented itself in this research that sought to try and overcome these issues.
+Although the research did not yield any significant improvements through translating the German to English as a means of extending the English dataset, it ultimately worsened the prediction results.
+We believe one of the main culprits is that the data for learning based on eight classes was insufficient.
+As can be seen in Equation \ref{eq:sentence-per-class} from Section \ref{sec:findings-and-implications-for-improving}, the models seemingly resorted to memorizing the sentence rather than learning the intricacies of the labeled classes.
+This inference arises that certain labels exhibited better F1 Score performance, primarily due to their greater availability in the training data.
 
@@
 While Google Cloud's translation service sufficed for this research, superior paid subscription-based translation services are available.
-With the help of more data and better translation more concrete answers to prediction improvement may be achieved.
+With the help of more data and better translation more concrete answers to improving predictions may be achieved.
 Secondly, another avenue to overcome the lack of data involves employing AI to generate sentences that are similar yet distinct from the original.
